# Functions in R

A function is a grouping of code. It helps you organize your code and reduce complexity.


## Bad Example

Look for examples of copy + pasted code in the same below.
What happens if we want to make an adjustment to all blocks of code?

``` {r}
library(tidyverse)

data('mpg')
t_mpg <- mpg


avg_cyl = mean(t_mpg$cyl)
rnd_avg_cyl <-  round(avg_cyl, 2)
text_avg_cyl <- paste('The average is ', rnd_avg_cyl) 
print(text_avg_cyl)


avg_hwy = mean(t_mpg$hwy)
rnd_avg_hwy <-  round(avg_hwy, 2)
text_avg_hwy <- paste('The average highway mpg is ', rnd_avg_hwy) 
print(text_avg_hwy)


avg_cty = mean(t_mpg$cty)
rnd_avg_cty <-  round(avg_cty, 2)
text_avg_cty <- paste('The average city mpg is ', rnd_avg_cty) 
print(text_avg_cty)


```


## Good Example

Let's break that code out into a separate function.

``` {r}
library(tidyverse)

data('mpg')
t_mpg <- mpg

find_avg <- function(text, field) {
  m = mean(t_mpg$cyl)
  rnd <-  round(m, 2)
  text <- paste('The average of', text, 'is', rnd) 
  return(text)
}

print(find_avg('highway mpg', t_mpg$hwy))
print(find_avg('city mpg', t_mpg$cty))


```

## Benefits

Here are some examples of some good functions.

``` {r}

# This is a good example of a function.
# It takes in a parameter, and returns the result.
# The scope of the function all stays inside of the function, with no
# contact with the outside items.
add_one <- function(x) {
  x1 <- x + 1
  return(x1)
}
y <- 0
print(add_one(y))


# Note scope! Variables inside of our code don't last after the
# function finishes running
add_one <- function(x) {
  temporary_variable <- x
  return(x + 1)
}
# The temporary_variable doesn't exist anymore!
add_one(1)


```

## Rules! 

There are some rules for creating and modifying functions.

``` {r}


# Don't update the parameters. Instead, return the result.
add_one_bad <- function(x) { 
  x <- x + 1
  return(x)
}
add_one_good <- function(x) { 
  x_new <- x + 1
  return(x_new)
}


# Don't have side effects.
#
# It is possible to break the connection between inside of the function
# and outside.
#
# To do so, use <<- instead of <-.  The latter just creates a new local
# variable inside of the block of code.
#
# In general, this is a bad idea.
y <- 0
add_to_y <- function() { 
  # This doesn't create a new variable, instead it updates y outside of the 
  # function block.
  y <<- y + 1 
}
add_to_y()


# Don't access variables outside of your function.
#
# This is legal, but a bad idea because it makes it harder
# to update and understand your code.
z <- 1
y <- 0

add_bad <- function(x) { return(x + z) }
add_bad(y)

add_good <- function(x1, x2) { return(x1 + x2) }
add_good(y, z)

```

