```{r, include=FALSE}

# Show code
knitr::opts_chunk$set(echo = TRUE)

# Add libraries
library(dplyr)
library(yarrr)
```

# dplyr

dplyr is an important tool for fixing data.   If you need further help,
 see [dplyr tutorial](https://dplyr.tidyverse.org/articles/dplyr.html)

## Why tibbles?

Data.frames are an older technology, and tibbles are their newer
version. Tibbles do not  automatically change data types. A data.frame guesses 
at times, but a  tibble must be told exactly what to do with incoming data. 
For example, a factor must be manually created. 
Tibbles also are better at printing.

## Converting from data.frame to tibble

If you've used data.frames, you may have used the approach below for subsetting.
The syntax for filtering rows/columns is pretty twitchy. dplyr is easier.

```{r, eval=FALSE, echo=TRUE }

# Create a data.frame
df <- data.frame(x = c(1, 2, 3, 4, 5), y = c("a", "b", "c", "d", "e"))


# Only include values with x>3
# Note that we use [rows_to_include, columns_to_include]
df[df$x > 3, ]

# Subset, finding a specific column. Return different types.
typeof(df[1])
typeof(df$x)
is.vector(df$x)
class(df[1])

```

## Creating a tibble

You will eventually load tibbles data from a file, but for now I will
usually include them in the file.

```{r, eval=FALSE, echo=TRUE }

# a tibble created from vectors.
t <- tibble(x = c(1, 2, 3, 4), y = c(5,  6,  7,  9), z = c("?"))

# You can subset a tibble and get a tibble as a result
t["x"]
t[c("x", "y")]

# Getting a vector requires double brackets.
t[["x"]]
class(t[["x"]])

```


```{r, eval=FALSE, echo=F }
# Application
# 
# Create a new tibble that looks like this:
#| Name | id | sales
#| Bob  | 1  | 123
#| Sam  | 2  | 848
s <- tibble(
  name = c("Bob", "Sam"), 
  id = c(1, 2), 
  sales = c(123, 456)
)

# Return just the name and sales as a tibble
s[c("name", "sales")]

# Show just sales as a vector
s[["sales"]]

```

## select

Select allows us to grab columns from a dataset. Usually there are a lot
more columns than we need.

```{r, eval=FALSE, echo=TRUE }

# Create a tibble from pirates (an older-style data frame)
class(pirates)
p <- tibble(pirates)
names(p)

# Simple reduced length tibble (not data.frame)
# Note that we start with the datasource, and then give each column.
p1 <- select(p, sex, age, college, beard.length, grogg)
names(p1)

# Select all EXCEPT certain columns
# You you can't mix this with other select options.
p2 <- select(p, -fav.pixar, -sword.type, -favorite.pirate)
names(p2)

# Select a range of columns (from:to)
p3 <- select(p, id:weight)
names(p3)

# You can combine ranges and individual fields.
p4 <- select(p, id, sex:weight, college:parrots)
names(p4)

# Use select and with to find all numeric columns
# The below finds all numeric columns and the sword.type text value.
p5 <- select(p, where(is.numeric), sword.type)

# See the docs for contains, ends_with, started_with, etc...

```


```{r, eval=FALSE, echo=F }
# Application

# Create a new variable called tp that creates a tibble from pirates
# Then make a new variable called tp... for each question.

tp <- tibble(pirates)
names(pirates)
# 1) Pirates id, sex, grogg, and eye patch
# 2) Exclude all text columns
# 3) Include all integer columns
# 4) Include all columns from sex to tchests.
tp1 <- select(tp, id, sex, grogg, eyepatch)
tp2 <- select(tp, -sex, -headband, -college, -favorite.pirate, -sword.type,
              fav.pixar)
tp3 <- select(tp, id)
tp4 <- select(tp, sex:tchests)
```


## filter

Removes rows (observations). Adding multiple items acts like an AND. You
can also use symbols and is &, not is !, or is \|.

```{r, eval=FALSE, echo=TRUE }
# Create a new tibble. Note that we use the single character name to keep
# our code shorter and easier to read. Only do this if the file is clearly
# focused on a single dataset.
p <- tibble(pirates)

# Count the number of males. Note that we can't use length, as that shows
# the numbers of columns.
p_nonmen <- filter(p, p$sex == "male")
nrow(p_nonmen)

# Use two conditions for an AND to find all non-male and non-female
p_other <- filter(p, p$sex != "female", p$sex != "male")
nrow(p_other)

# Use | (or) to find male and females
p_male_and_female <- filter(p, p$sex == "male" | p$sex == "female")
nrow(p_male_and_female)

# Make row 1 sex column have an NA for testing purposes.
p[1, "sex"] <- NA

# Remove NA values, which can be an issue in some datasets.
# You can't test for sex == NA, or sex == "NA", as all NA are unique.
# Instead, use the is.na function.
p_no_nas <- filter(p, !is.na(p$sex))
nrow(p_no_nas)

```


```{r, eval=FALSE, echo=F }
# Application

# Create a new dataset called tp from pirates for this section.
tp <- tibble(pirates)

# Remove the sex from rows 3 and 5, replacing with NA
tp[c(3, 5), "sex"] <- NA

# Find pirates with a weight over 90
tp1 <- filter(tp, weight > 90)

# Find pirates with a weight over 90 and under 100
tp2 <- filter(tp, weight > 90 & weight < 100)

# Find Women pirates
tp3 <- filter(tp, sex == "female")

# Find female pirates with a weight over 90
tp4 <- filter(tp, sex == "female", weight > 90)

# Find pirates with a sex of NA
tp5 <- filter(tp, is.na(sex))

# Find pirates without a sex of NA
tp6 <- filter(tp, !is.na(sex))
```


## mutate

Adds a new column. This really helps us manipulate data for later analysis.

```{r, eval=FALSE, echo=TRUE }

# Create a new tibble with fewer columns
# Note that we are updating this same tibble instead of creating a new one
# each time.
p <- select(tibble(pirates), id, sex, age)

# Create new columns for is_male and is_female
# Each column will be either TRUE or FALSE.
p <- mutate(p, is_male = sex == "male")
p <- mutate(p, is_female = sex == "female")

# Use ifelse to use other values, such as 1 or 0, for a logical test.
p <- mutate(p, is_male01 = ifelse(sex == "male", 1, 0))

```


```{r, eval=FALSE, echo=F }
# Application


# Create a new tibble with ID, Tattoos, and head band columns.
# Update this tibble in this section with mutate commands.
t <- select(tibble(pirates), id, tattoos, headband)

# Add a new TRUE/FALSE column for many tattoos over 10.
t <- mutate(t, many.tattoos = tattoos > 10)

# Add a new 1/0 column for many tattoos over 10
t <- mutate(t, many.tattoos01 = ifelse(tattoos > 10, 1, 0))

# Add a new "yes" or "no" column for many tattoos over 10
t <- mutate(t, many.tattoosYN = ifelse(tattoos > 10, "yes", "no"))

# Use print to printing the resulting dataset.
print(t)
```

## arrange

Re-arrange (or sort data)

```{r, eval=FALSE, echo=TRUE }

# Create a dataset from pirates, select ID, Age, and Grogg, and then sort.
# Note that we can use desc(fieldname) to sort in reverse order.
p <- arrange(select(tibble(pirates), id, age, grogg), age, desc(grogg))
print(p)
```


```{r, eval=FALSE, echo=F }
# Application

# Create a new tibble p with id, sex, college, and tattoos.
# Then filter to only have CCCC members with over 10 tattoos.
# Create new fields called is_male with the values 1/0, and many_tattoos
#   with the values of "Over 10", and "Other"
# Order by tattoos, and sex in reverse order.
# Print the table.
p <- tibble(pirates)
p <- select(p, id, sex, college, tattoos)
p <- filter(p, college == "CCCC" & tattoos > 10)
p <- mutate(p, is_male = ifelse(sex == "male", 1, 0))
p <- mutate(p, many_tattoos = ifelse(tattoos > 10, "Over 10", "Other"))
p <- arrange(p, tattoos, desc(sex))

print(p)
```

## group_by and summarize

The function group_by is very similar to the SQL command, but in a
passive way. Normally, you would expect it to immediate collapse data
(similar to a pivot table). However, it is better thought of giving an
annotation to the data that will then be used by the summarize command.

Calculates summary statistics for a field. Use mean, min, max, sum, n,  
n_distinct.

```{r, eval=FALSE, echo=TRUE }

# Create a tibble and group by college
p <- group_by(tibble(pirates), college)


# Find the median, length (count), max, and min age by college
summarise(p,
          median_age = median(age),
          length_age = length(age),
          max_age = max(age),
          min_age = min(age),
          n_of_rows = n(),
          n_of_distinct_ages = n_distinct(age)
          )

```


```{r, eval=FALSE, echo=FALSE }
# Application

# Create a tibble with id, sword.type, sword.time, and favorite.pirate
p <- select(tibble(pirates), id, sword.type, sword.time, favorite.pirate)

# Remove all pirates with a favorite of Hook, or who have an ID under 500
p <- filter(p, favorite.pirate == "Hook" | id < 500)

# Add a new 1/0 sword field that tests to see if the sword is a cutlass or not.
p <- mutate(p, cutlass01 = ifelse(sword.type == "cutlass", 1, 0))

# Group by sword.type
p <- group_by(p, sword.type)

# Show the average sword.time by sword.type
summarise(p, mean_sword.time = median(sword.time))

# Show the average and standard deviation of sword.time by favorite pirate
# hint: use group by
p <- group_by(p, favorite.pirate)
summarise(p,
          mean_sword.time = median(sword.time),
          sd_word.time = sd(sword.time))
```



## joins

Similar to SQL, we will use joins in dplyr. Note that this uses very weird 
syntax. There are multiple ways to specify the join, but try to always use
the by = c("field_1" = "field_2") version. This is the most 
consistently useful approach.


```{r, eval=FALSE, echo=TRUE }

# Create sample tibbles.
checks <- tibble(
  check_id = c(1, 2, 3, 4, 5),
  desc = c("Cheese", "Bread", "Milk", "Mustard", "Milk"),
  vendor_fk = c(2, 2, 2, 3, 5),
  amount = c(10, NA, 30, 40, 50)
)

vendors <- tibble(
  vendor_pk = c(2, 3, 4, 9),
  desc = c("Bob", "Sarah", "Tim", "Andrew"),
)

# Create a left, right, and inner join for checks and vendors.
# Note the different results
left_join(checks, vendors, by = c("vendor_fk" = "vendor_pk"))
right_join(checks, vendors, by = c("vendor_fk" = "vendor_pk"))
inner_join(checks, vendors, by = c("vendor_fk" = "vendor_pk"))

# Note that we can have issues with NA values.
sum(checks$amount)

# use na.rm = TRUE to remove.
sum(checks$amount, na.rm = TRUE)

```

## pipe operator

We can put these all together using the %\>% (pipe) operator. This joins
multiple results without needing to re-type stuff.

Note that all dplyr functions use the tibble as the first argument.
Rather than nesting or using intermediate variables, %\>% allows us to
rewrite them.

x %\>% f(y) turns into f(x, y)

Read left-to-right, top-to-bottom.

You can type with Control+Shift+M

```{r, eval=FALSE, echo=F }
# Application

# We can save each intermediate step.
p <- select(pirates, sex, age, college, beard.length, grogg)
p <- mutate(p, is_male = p$sex == "male")
p <- mutate(p, is_female = p$sex == "female")
p <- arrange(p, p$age)


# Or just use the pipe
p <- tibble(pirates) %>%
  select(sex, age, college, beard.length, grogg) %>%
  mutate(isCCCC = college == "CCCC") %>%
  mutate(is_male = sex == "male") %>%
  mutate(is_female = sex == "female") %>%
  arrange(age)

print(p)
```
