```{r, include=FALSE}

# Add libraries
library(tidyverse)

```


# ggplot 1

ggplot is a super useful graphics tool. See its website at [dplyr
tutorial](https://ggplot2.tidyverse.org/)

There is also a very nice [cheat cheat by Afshine Amidi & Shervine] Amidi(https://github.com/shervinea/mit-15-003-data-science-tools/blob/master/en/study-guide-data-visualization-with-r.pdf)

Here is a great systematic introduction 
[Elegant Graphics for Data Analysis](https://ggplot2-book.org/)

Here is a list of example charts
[R Graphics Cookbook](https://r-graphics.org/)


## Why ggplot?

R has a number of built-in graphical functions. However, they are
limited in customizability, and all work a little differently. ggplot is
a consistent approach that allows you to build very elegant plots.

## Creating a plot

You will eventually load tibbles data from a file, but for now I will
usually include them in the file. We are going to use a sample dataset
called mpg.

First, call `ggplot` to start your plot. It's easiest to give it the data
at this point.

Note that nothing happens. This is because we need to draw stuff on the
chart.

```{r, eval=T, echo=TRUE }
ggplot(data = mpg)
```

## Geoms for 1-2 variables 

There are different types of geoms:

- Individual: `geom_area`, `geom_col`, `geom_line`, `geom_point`
- Collective: `geom_boxplot`, `geom_histogram`
- Maps: `geom_sf`


### geom_bar

A bar plot will show the *count* of items at every point in an axis
(either x or y).

Note that we use the + to join it to the ggplot.

We need to add a mapping argument to geom_bar, and will use an aesthetic
(aes). This allows us to define which parts of the chart will map to our
dataset.

Any properties that don't map to the `aes` (data set) need to be placed
outside of the mapping argument. See examples below.

Here is a minimal example

```{r, eval=T, echo=TRUE }
# Minimal example
ggplot(data = mpg) +
  geom_bar(mapping = aes(y = hwy))
```

With some options:

- x/y: use either x or y, but not both. The other will be count of rows.
- fill: text of color.
- color: border (text color) 

```{r, eval=T, echo=TRUE }

# Example with color/fill
ggplot(data = mpg) +
  geom_bar(mapping = aes(x = hwy),
           color = 'blue',
           fill = 'black')
```


### geom_smooth

The smooth geom shows a smoothed line. It shows the *average* value at each
point, instead of the bar (which defaults to count).

Note that we need both an X and a Y (numeric).

```{r, eval=T, echo=TRUE }
# Show with error bars
ggplot(data = mpg) +
  geom_smooth(mapping = aes(y = hwy, x = cty))

# Full example
# method: loess is default for small items. Can also use lm (linear reg)
# color: line color
# linewidth: width for the bar (number)
# se: False to use no error bars (confidence interval around smooth)
# na.rm: T to remove na values with no error
# stat: default is smooth (shows a smoothed line). To show all, use identity.
ggplot(data = mpg) +
  geom_smooth(mapping = aes(y = hwy, x = cty), 
              method = 'lm',
              na.rm = TRUE,
              stat = 'identity',
              se = FALSE,
              color = 'green',
              linewidth = 3)

```

### geom_line

This is identical to `geom_smooth`, but defaults to showing the stat as 
identity, instead of smooth.

```{r, eval=T, echo=TRUE }
# Show with error bars
ggplot(data = mpg) +
  geom_line(mapping = aes(y = hwy, x = cty))
```

### geom_point

This shows a separate dot for each row in the dataset (identity) 

```{r, eval=T, echo=TRUE }

# Basic example
#   x and y: can be either categorical or numerical data
ggplot(data = mpg) +
  geom_point(mapping = aes(y = hwy, x = class))

# We have a lot of properties that can be mapped
#   shape: best discrete, (suggested max of 6 different items)
#   size: continuous value
#   alpha: continuous values
#   color: either continuous or discrete
ggplot(data = mpg) +
  geom_point(mapping = aes(y = hwy, 
                           x = manufacturer,
                           shape = class,
                           size = cyl,
                           color = cty,
                           alpha = displ))

# Note that this can also be set manually
#   shape: Number, 1-20
#   size: Number 
#   alpha: Number 0-1
#   color: Text, such as "black" or a hex code "#2211aa"
ggplot(data = mpg) +
  geom_point(mapping = aes(y = hwy, 
                           x = manufacturer),
             shape = 2,
             size = 3,
             color = '#00bb00',
             alpha = 0.1)



```



#### jitter

Another useful function combined with point is `jitter.` This will slightly
adjust the values given, making it easier to see overplotted data.

Use in combination with a low alpha value to see density of points on your plot.

```{r, eval=T, echo=TRUE }

# jitter takes in a variable, and adjusts it by the factor
# Note that it only works with continuous data
ggplot(data = mpg) +
  geom_point(mapping = aes(y = jitter(hwy, factor = 1.5), 
                           x = manufacturer),
             
             alpha = 0.1)
```

## Geoms for distributions

### geom_histogram

Show the count of items using bins

```{r, eval=T, echo=TRUE }
ggplot(data = mpg) +
  geom_histogram(mapping = aes(x = hwy),
                 binwidth = 5)
```


### geom_density

Similar to a histogram, but you can layer different lines on top of each other.

```{r, eval=T, echo=TRUE }
ggplot(data = mpg) +
  geom_density(mapping = aes(x =hwy, color = class, fill = class),
               alpha = 0.2)
```

### geom_boxplot

Show a boxplot for a series of discrete values (usually text)

```{r, eval=T, echo=TRUE }
ggplot(data = mpg) +
  geom_boxplot(mapping = aes(x = class, y = hwy))
```





## Facets

We can create multiple plots using facets

### facet_wrap

Show a separate plot for each categorical data

```{r, eval=T, echo=TRUE }

# Small multiples (facets) are a great technique for exploring categorical data
# and its relationship with continuous data.
#
# Note the ~ symbol.  ~ (tilde) is commonly use in lm (regression) to
# show [dependent variable] ~ [independent variable] For now, just
# remember to keep the left blank and put your category variable on the right.
# 
# nrow says how many plots to put per row
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_wrap( ~ class, nrow = 3)

```

### facet_grid

Show a grid of charts, using two continuous variables.

```{r, eval=T, echo=TRUE }

# Small multiples (facets) can be used with two separate variables.
# Use facet_grid instead of facet_wrap.
# Use categorical variables.
# Note the ~  uses drv (type of drive-train) and cyl (cylinders)
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(cyl ~ drv)

```








## Dynamic plots

Plotly can be used to make a plot more dynamic. See its [website](https://plotly.com/ggplot2) for more information.