# Special Topics


## Date

Date/times in R require some special consideration.
They are a distinct type, and require the as.Date function to turn from text (i.e., '01-01-2022' into a date).

``` {r}

date_as_text = '2001-04-03'
date_as_date = as.Date(date_as_text)

print(class(date_as_date))
```

## Format function

We can output dates in different styles with the `format` command: 

- `%d` number day as `19`
- `%a` text weekday as `Sun` (or `%A` for `Sunday`)
- `%m` number month as `04`
- `%b` text month as `Feb` (or `%B` as `February`)
- `%y` year as `14` (`%Y` as `2014`)

``` {r}

d <- as.Date('2001-07-01')

vector_of_date_as_text_values <- c(
  format(d, format = '%d %m %y'),
  format(d, format = '%d-%m-%y'),
  format(d, format = '%B, %Y'),
  format(d, format = '%a (%d of %b)')
)
print( vector_of_date_as_text_values)
```

## Lubridate

We typically use lubridate in this class to simplify date/time handling. The biggest issue to keep straight is if an item is a date or a date-time. See below for examples.


``` {r}
library(lubridate)


d <- ymd('2023-02-03')
dt <- ymd_hm('2023-02-03 12:23')


```


## Cool articles on time

How do we deal with leap dates? What about leap seconds?
 
- [Leap seconds](https://arstechnica.com/science/2023/11/leap-seconds-could-become-leap-minutes-despite-pushback-from-russians-vatican/)