---
title: EDA
output: 
  html_document:
    toc: true
    theme: united
---

# EDA in R - Introduction

This corresponds to the DataCamp *Exploratory Data Analysis in R* course.

It helps us understand how multiple variables are related to each other.

# Concepts

Marginal: the chance that of an event happening. Example, what are the chances of a heads on a coin flip, given that we had just had a head. It is still 50%. 

Conditional: If A, then the proportion of B changes
As an example: If a person is man, then they are more likely to be tall.

# Categorical data

Process:

- Find unique values for a field
  - low-count items as needed
  - Use both count and proportions
- Remove bad categories
 

## table

Print raw count of two categorical variables

```{r echo=TRUE}
library(tidyverse)

t <- tibble(x = c('Bob', 'Bob', 'Bob', 'Sue', 'Sue'), 
            z = c('a', 'b', 'a', 'a', 'b'))


# Print normal table
table(t$x, t$z)
```

## Proportion table

Show the counts as proportions

```{r echo=TRUE}
library(tidyverse)

t <- tibble(x = c('Bob', 'Bob', 'Bob', 'Sue', 'Sue'), 
            z = c('a', 'b', 'a', 'a', 'b'))

# Change printout to simplify output
options(scipen = 999, digits = 3)

# Print table of % (of total)
#   For 2nd argument, 
#     NA: % of total
#     1: % of row
#     2: % of column
prop.table(table(t$x, t$z), 2)
```

## Charts

Good chart designs:

- Stacked bar chart
- Faceted bar charts

# Distribution data

We generally need a chart that shows a range of data.

- Point
  - Points are good for smaller sets (n < 100), but have issues with overplotting.
- Histogram
  - Very good for one variable
  - Need facets to add in categorical variable
- Density plot
- Boxplot
  - Great for large numbers of categorical variables
  
  
# Plots for 3 variables

Ensure that we have enough data points for each combination of variables.

High dimension plots can easily be created through the use of facets. But, dotplots are also a good approach.


# Measures 

## Centrality

Process:

- Create a group by
- Summarize iwth mean, median, mode

## Variability

var:

- Measure distance between each item and mean
- Square it, to create all positives and increase weight of bigger differences.
- Divide by the count
- End up with units^2

sd: same as var, but square root (original units)

summary gets quartiles

IQR: inter-quartile range.



# Shape and Transformations


Modality: 
- Unimodal: one point
- Bimodal: two
- Multimodal: 3+

Skew: left or right (where the long tail is!)

Transform by applying a log transformation.

# Outliers

Solutions:

- Create a new variable to identify outliers and plot
- Filter out outliers


# Zero-inflation strategy

We often have issues with variables having a lot of zeros. We can convert that into a yes/no variable.
