---
title: Regression 1
output: 
  html_document:
    toc: true
    theme: united
---

This tutorial aligns with DataCamp's *Introduction to Regression in R* tutorial.

# Linear Regression {.tabset}

Linear Regression is an incredibly important tool. It uses multiple numerical independent variables to predict a single output/dependent variable.

The output variable is generally a *continuous* numerical variable.

Outcomes:

- Define the goal of linear regression
- Describe when using a linear regression model is useful
- Describe dependent and independent variables
- Describe linear and non-linear relationships
- Transform inputs to meet various requirements

Terms:

- Coefficients, p-values
- Residuals, sse (sum of squares error, or a loss function)
- Bias / intercept
- R^2

Help: 


- [Linear Regression in 3 minutes](https://www.youtube.com/watch?v=3dhcmeOTZ_Q)
- [Multiple regression in R](https://www.youtube.com/watch?v=hokALdIst8k)
- [Multiple regression in theory](https://www.youtube.com/watch?v=zITIFTsivN8)
- [Multiple linear regression tutorial](https://bradleyboehmke.github.io/HOML/linear-regression.html)
- [Should we use a p<0.05 in science?](https://www.tandfonline.com/doi/full/10.1080/00031305.2019.1583913)

## Minimal Example

```{r echo=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)

t_data <- tibble(
  from = c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10),
  to = c(10, 10, 8, 8, 7, 6, 5, 4, 2, 1),
  extra = c(0, 1, 0, 1, 0, 1, 0, 1, 1, 1)
)

# Quick viz of data showing the relationship.
pairs( t_data )

# Create a linear regression model predicting `to` with `from` and using the entire dataset.
model = lm(to ~ from, data = t_data)

# Show a summary of the linear model. Pay attention to these key items:
#   Residuals = errors in each row between actual & predicted
#   Std Error = the squared difference between
#      the predicted and actual values. 
#   Coefficients: 
#     Estimate: value of the change in input to output
#     Std. Error: averages squared diff between prediction / actual
#     p Value: probability of the estimate being a result of random chance
#   Residual standard error: the overall avg difference between actual
#     and predicted for the entire model. 
#   Adj R^2,  the % of variation explained by the model
summary(model)
```

## Full example

```{r echo=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)

t_data <- tibble(
  from = c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10),
  to = c(10, 10, 8, 8, 7, 6, 5, 4, 2, 1)
)

# Quick viz of data showing the relationship. Look for any obvious correlation values.
# We cant to avoid any perfectly correlated data, as that will cause problems in our regression.
pairs( t_data )

# Show above, but in numbers.
cor( t_data, use = 'pairwise.complete.obs' )


# Create a linear regression model predicting `to` with `from` and using the train dataset.
model = lm(to ~ from, data = t_data)

# Show a summary of the linear model. Pay attention to these key items:
#   Residuals = errors in each row between actual & predicted
#   Std Error = the squared difference between
#      the predicted and actual values. 
#   Coefficients: 
#     Estimate: value of the change in input to output
#     Std. Error: averages squared diff between prediction / actual
#     p Value: probability of the estimate being a result of random chance
#   Residual standard error: the overall avg difference between actual
#     and predicted for the entire model. 
#   Adj R^2,  the % of variation explained by the model
summary(model)

# Show a plot of residuals / errors
hist(model$residuals)

# If we plug in a 95% confidence interval, how accurate are we?
confint(model, level = 0.95)


# Now look at how the data performs on the data.
# This function will take in a tibble and a model, test it,
#   and calculate the resulting average error per item.
rmse <- function( m, tibble, dependent_variable ) {
  results <- predict(m, tibble)
  errors <- results - dependent_variable
  return( sqrt(mean(errors^2, na.rm = TRUE)))
}

rmse(model, t_data, t_data$to)
```

# Preprocessing  {.tabset}

We have some clean-up to do with variables used in regression.

## Low-variance variables

Look for values that have low variation and remove them from the dataset.

```{r echo=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)
library(caret)

t_data <- tibble(
  from = c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10),
  to = c(10, 10, 8, 8, 7, 6, 5, 4, 2, 1),
  group_a = c(1, 1, 1, 1, 1, 1, 2, 1, 1, 1),
  group_b = c(3, 3, 1, 1, 1, 1, 2, 1, 2, 5),
)

# saveMetrics give us output description. 
#   freqRatio: top most frequent item / 2nd top most freq item. Goal is to be close to 1
#   percentUnique: unique(items) / total items
#
# Output
#   freqRatio: 
#     from and to are 1.0, meaning that the most frequent value is just as likely
#       as the next-most likely value. 
#     However, group_a is 9.0, meaning that the most likely value  (1) is 9 times as common 
#       as the next most likely (2). Remove from our model!
#     Group_b is better, with the most likely value (1) being 2.5 times as likely as the 
#       next most (2 or 3)
#   
nearZeroVar(t_data, saveMetrics = TRUE)


```


## Remove perfect correlations

Look for values that are perfectly-correlated with each other, and remove all but *one*.

```{r echo=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggcorrplot)

t_data <- tibble(
  from = c(1, 2, 3, 4, 5, 6, 7, 8, NA, 10),
  to = c(10, 10, 8, 8, 7, 6, 5, 4, 2, 1),
  is_major = c(1, 1, 1, 1, 1, 1, 0, 0, 0, NA),
  is_not_major = c(0, 0, 0, 0, 0, 0, 1, 1, 1, NA),
  from_x_10 = from * 10
)

# Pairs shows a visual of the relationship in our model
# Look for numerical values in a straight line
pairs( t_data)

# Cor returns the correlation. Look for values near 1.0 or -1.0
# use allows us to use more data, ignoring pairs with a NA 
cor( t_data, use = 'pairwise.complete.obs' )

# Visual of above 
ggcorrplot::ggcorrplot(cor( t_data, use = 'pairwise.complete.obs' ))
```


## Remove outliers

Look for values with outlier information

```{r echo=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggcorrplot)

t_data <- tibble(
  from = c(1, 2, 3, 4, 5, 6, 7, 8, NA, 10),
  to = c(10, 10, 8, 8, 7, 6, 5, 4, 2, 1),
  absences = c(0, 0, 0, 23, 0, 0, 2, 3, 1, 1)
)

# Hist is a quick function to display outlier values.
# As a general rule, we may want to remove values that are >1.5 the 
# interquartile range (difference between 50% and 75% values). However, this is a
# judgement call.
hist( t_data$absences)
summary( t_data$absences)

# To cleanup, you can take 3 approaches:
clean_t_data <- t_data %>% 
  # option a: remove rows
  filter(absences < 50) %>%
  # option b: turn into a yes/no 1/0 variable
  mutate(absences_excessive = ifelse(absences > 5, 1, 0)) %>% 
  # option c: cap the field
  mutate(absences_capped = ifelse(absences > 5, 5, absences))

print(clean_t_data)
```



## Add dummy variables

Change text fields into 1/0 values.

```{r echo=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggcorrplot)

t_data <- tibble(
  from = c(1, 2, 3, 4, 5, 6, 7, 8, NA, 10),
  to = c(10, 10, 8, 8, 7, 6, 5, 4, 2, 1),
  major = c("Y", "Y", "Y", "Y", "Y", "Y", "N", "N", "N", NA)
)

table( t_data$major)

# To cleanup, use ifelse
clean_t_data <- t_data %>% 
  mutate(major01 = ifelse(major == 'Y', 1, 0))

print(clean_t_data)
```


## Collapse fields

You may also want to join text categories into smaller groups

```{r echo=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggcorrplot)

t_data <- tibble(
  from = c(1, 2, 3, 4, 5, 6, 7, 8, NA, 10),
  to = c(10, 10, 8, 8, 7, 6, 5, 4, 2, 1),
  major = c("Act", "Act", "Fin", "Act", "Marketing", "Act", "Act", "Act", "Act", "Act")
)

table( t_data$major)

# To cleanup, use ifelse
clean_t_data <- t_data %>% 
  mutate(major_not_accounting01 = ifelse(major == 'Fin' | major == "Marketing", 1, 0))

print(clean_t_data)
```
