---
title: dplyr
output: 
  html_document:
    toc: true
    theme: united
---

dplyr is an important tool for fixing data.   If you need further help,
 see [dplyr tutorial](https://dplyr.tidyverse.org/articles/dplyr.html)

These notes correspond to Datacamp's *Introduction to the tidyverse*.


# Why tibbles?

Data.frames are an older technology, and tibbles are their newer
version. Tibbles do not  automatically change data types. A data.frame guesses 
at times, but a  tibble must be told exactly what to do with incoming data. 
For example, a factor must be manually created. 
Tibbles also are better at printing.

# Converting from data.frame to tibble

If you've used data.frames, you may have used the approach below for subsetting.
The syntax for filtering rows/columns is pretty twitchy. dplyr is easier.

```{r, eval=F, echo=T, message=FALSE }
library(tidyverse)

# Create a data.frame
df <- data.frame(x = c(1, 2, 3, 4, 5), y = c("a", "b", "c", "d", "e"))

# Only include values with x>3
# Note that we use [rows_to_include, columns_to_include]
df[df$x > 3, ]

# Subset, finding a specific column. Return different types.
typeof(df[1])
typeof(df$x)
is.vector(df$x)
class(df[1])

```

# Creating a tibble

You will eventually load tibbles data from a file, but for now I will
usually include them in the file.

```{r, eval=F, echo=TRUE, message=FALSE }
library(tidyverse)

# a tibble created from vectors.
# Note that either they must have the same length,
# or a length of zero (which are automatically expanded)
t <- tibble(
  x = c(1, 2, 3, 4),
  y = c(5,  6,  7,  9),
  z = c("?")
)

# You can subset a tibble and get a tibble as a result
t["x"]
t[c("x", "y")]

# Getting a vector requires double brackets.
t[["x"]]

# You can verify the output by using the class function.
class(t[["x"]])

```

# dplyr commands

## select {.tabset}

Select allows us to choose columns. It helps when we load a large dataset with more columns than we need.

### Columns by name

Pick columns by passing their field names, with each separated by a comma.

```{r, eval=T, echo=TRUE, message=FALSE }
library(tidyverse)

t <- tibble(
  names = c("Bob", "Sarah", "Tim", "John"),
  ages = c(23,  61,  17,  9),
  id = c(1, 2, 3, 4)
)

t_names <- t %>% 
  select(names, ages)

print(t_names)
```


### Remove columns by name

Remove specific columns by placing a `-` in front of each.

```{r, eval=T, echo=TRUE, message=FALSE }
library(tidyverse)

t <- tibble(
  names = c("Bob", "Sarah", "Tim", "John"),
  ages = c(23,  61,  17,  9),
  id = c(1, 2, 3, 4),
  null_field = c(NA, NA, NA, NA)
)

t_no_id_or_null <- t %>% 
  select(-id, -null_field)

print(t_no_id_or_null)
```



### Add columns by name array

You can include a list of columns with `start_column:end_column`.

```{r, eval=T, echo=TRUE, message=FALSE }
library(tidyverse)

t <- tibble(
  extra = 1:4,
  name = c("Bob", "Sarah", "Tim", "John"),
  ages = c(23,  61,  17,  9),
  id = c(1, 2, 3, 4),
  null_field = c(NA, NA, NA, NA)
)

t_name_to_id <- t %>% 
  select(name:id)

print(t_name_to_id)
```


### All numeric columns

You can also use the `where` function pick only fields meeting a condition. 


```{r, eval=T, echo=TRUE, message=FALSE }
library(tidyverse)

t <- tibble(
  extra = 1:4,
  name = c("Bob", "Sarah", "Tim", "John"),
  ages = c(23,  61,  17,  9),
  id = c(1, 2, 3, 4),
  null_field = c(NA, NA, NA, NA)
)

t_numbers <- t %>% 
  select(where(is.numeric))

print(t_numbers)
```


### Helper functions

We can use helper functions to select fields:

- `starts_with()`
- `ends_with()`
- `contains()`
- `matches()`

See the docs for other options.

```{r, eval=T, echo=TRUE, message=FALSE }
library(tidyverse)

t <- tibble(
  id = c(1, 2, 3),
  population_start = c(100, 200, 300),
  population_middle = c(110, 150, 200),
  population_end = c(120, 80, 20),
  not_needed_field = c(NA, "hey", "umn")
)

t_smaller <- t %>% 
  select(starts_with('population'), id)

print(t_smaller)
```


## filter {.tabset}

Removes rows (observations). Uses the symbols:

- `&` as *and*
- `!` as *or*
- `|` as *not*

### Basic example

```{r, eval=T, echo=TRUE, message=FALSE }
library(tidyverse)

t <- tibble(sex = c("man", "woman", "man", "woman", "man"))

t_men_only <- t %>%
  filter(sex == "man")

print(t_men_only)
```

### And

You can combine multiple logical tests with 3 different approaches

```{r, eval=T, echo=TRUE, message=FALSE }
library(tidyverse)
t <- tibble(sex = c("man", "woman", "man", "woman", "man"),
            age = c(10, 20, 30, 40, 50))


# Use one the following:

# Option 1: Use & between each test
t_men_over_30 <- t %>%
  filter(sex == "man" & age > 30)

# Option 2: Separate each test with a comma
t_men_over_30 <- t %>%
  filter(sex == "man",  age > 30)

# Option 3: Use a separate filter for each logical test
t_men_over_30 <- t %>%
  filter(sex == "man") %>%
  filter(age > 30)

print(t_men_over_30)
```

### Or

The `|` symbol is used when we want to satisfy *at least one* of several logical tests.

```{r, eval=T, echo=TRUE, message=FALSE }
library(tidyverse)
t <- tibble(sex = c("man", "woman", "man", "woman", "man"),
            age = c(10, 20, 30, 40, 50))

t_men_or_those_over_30 <- t %>%
  filter(sex == "man" | age > 30)

print(t_men_or_those_over_30)
```


### Not

Use the `!` symbol to test for anything *not true* (or *false*).

```{r, eval=T, echo=TRUE, message=FALSE }
library(tidyverse)
t <- tibble(sex = c("man", "woman", "man", "woman", "man"),
            age = c(10, 20, 30, 40, 50))

t_not_age_30 <- t %>%
  filter(!age == 30)

print(t_not_age_30)
```

### Test for NA values

Test for `NA` values with `is.na()`. You can remove `NA` values  by filtering for `!is.na(column_name)`.

Each `NA` value is unique. So, testing for `column_name == NA` will always be `FALSE`.

```{r, eval=T, echo=TRUE, message=FALSE }
library(tidyverse)
t <- tibble(sex = c("man", "woman", NA))

t_not_na <- t %>%
  filter(!is.na(sex))

print(t_not_na)
```


### Filter by list

We can use the `%in%` command to find values in a vector.

```{r, eval=T, echo=TRUE, message=FALSE }
library(tidyverse)
t <- tibble(tool = c("driver", "saw", "nail", "hammer"))

t2 <- t %>%
  filter(tool %in% c("nail", "hammer"))

print(t2)
```



## mutate {.tabset}

Adds a new column or changes an existing one.

### Change values in a column 

```{r, eval=T, echo=TRUE, message=FALSE }
library(tidyverse)
# population in millions
cities <- tibble(population = c(10, 20, 15, 2, 3))

# Change population from millions to thousands
cities <- cities %>% 
  mutate(population = population * 1000)

print(cities)
```

### Change value to ratio

We can change a value into a percentage (or ratio) by using `sum()` and then dividing a field by that value.

```{r, eval=T, echo=TRUE, message=FALSE }
library(tidyverse)
# population in millions
cities <- tibble(population = c(10, 20, 15, 2, 3))

total_population <- sum(cities$population)

# Change population from millions to thousands
cities <- cities %>% 
  mutate(population = population / total_population)

print(cities)
```

### New Boolean column 

```{r, eval=T, echo=TRUE, message=FALSE }
library(tidyverse)
t <- tibble(sex = c("man", "woman", "man", "woman", "man"))

t_is_man <- t %>% 
  mutate(is_man = sex == "man")

print(t_is_man)
```

### ifelse returning 1/0

We typically use `ifelse()` to create new columns. The first argument will be a *logical test*. If the test is `TRUE`, the function will return its second argument. If the test is `FALSE`, it will return the third argument.

```{r, eval=T, echo=TRUE, message=FALSE }
library(tidyverse)
t <- tibble(sex = c("man", "woman", "man", "woman", "man"))

# Function:
#    ifelse(*logical_test*, *result_if_true*, *result_if_false*)
# 
# Below gives a 1 for males, and a 0 for all others
# 
# Note that we have to end with two parenthesis, one for the ifelse and one for the mutate.
t_with_is_man <- t %>% 
  mutate(is_man01 = ifelse(sex == "man", 1, 0) )

print(t_with_is_man)
```


### ifelse returning string

This example shows `ifelse()` returning a string.

```{r, eval=T, echo=TRUE, message=FALSE }
library(tidyverse)
t <- tibble(sex = c("man", "woman", "man", "woman", "man"))

t_with_m <- t %>% 
  mutate(s = ifelse(sex == "man", "m", "w") )

print(t_with_m)
```


### Multiple new columns

We can add multiple columns with a single mutate.
Below adds two new columns and then modifies the `sex` field.

```{r, eval=T, echo=TRUE, message=FALSE }
library(tidyverse)
t <- tibble(sex = c("Man", "Woman", "MAN", "wOMan", "man"))

t2 <- t %>% 
  mutate(sex_capitalized = str_to_upper(sex),
         sex_lower = str_to_lower(sex),
         sex = str_to_title(sex))

print(t2)
```


## arrange {.tabset}

The `arrange` functions sorts data.

### Ascending  

```{r, eval=T, echo=TRUE, message=FALSE }
library(tidyverse)
t <- tibble(sex = c("man", "woman", "man", "woman", "man"),
            age = c(20, 30, 40, 30, 50))

t_sorted <- t %>% 
  arrange(sex)

print(t_sorted)
```


### Ascending with two or more columns 

```{r, eval=T, echo=TRUE, message=FALSE }
library(tidyverse)
t <- tibble(sex = c("man", "woman", "man", "woman", "man"),
            age = c(20, 30, 40, 30, 50))

t_sorted <- t %>% 
  arrange(sex, age)

print(t_sorted)
```

### Descending order 

Wrap a field with `desc(field_name)` to reverse sort (z-a).

```{r, eval=T, echo=TRUE, message=FALSE }
library(tidyverse)
t <- tibble(sex = c("man", "woman", "man", "woman", "man"),
            age = c(20, 30, 40, 30, 50))

t_sorted <- t %>% 
  arrange(desc(sex), desc(age))

print(t_sorted)
```


## group_by and summarize {.tabset}

These two command are similar to SQL's `GROUP BY` or Excel's *Pivot Table*.

The `group_by` command is used first. It doesn't change the data, but instead creates a note on the tibble. 

The `summarize` command is used second. It reads the note on the tibble set by `group_by` to find which fields to include in the results. It then performs some type of aggregation (such as sum or count). The results are a new tibble.


### Summarize without group_by

`summarize()` is similar to mutate, in that both create a new column. But, each new field must have an aggregation function on it, such as `sum()`.

```{r, eval=T, echo=TRUE, message=F }
library(tidyverse)

t <- tibble(
  region = c('a', 'b', 'c', 'a', 'a', 'b'),
  sales = c(0, 10, 5, 3, 2, 1)
)

# Show the sum of dales by region
t_sum <- t %>% 
  summarise(sum_of_sales = sum(sales))

print(t_sum)
```

### Summarize with group_by

While `summarize` is helpful, we normally want to break the results down by some of the fields.

Start by using `group_by`, which includes region in our results. Then use `summarize()` to show which field we want to aggregate.

`summarize()` is similar to mutate, in that both create a new column.

```{r, eval=T, echo=TRUE, message=F }
library(tidyverse)

t <- tibble(
  region = c('a', 'b', 'c', 'a', 'a', 'b'),
  sales = c(0, 10, 5, 3, 2, 1)
)

# Show the sum of dales by region
t_region_sum <- t %>% 
  group_by(region) %>% 
  summarise(sum_of_sales = sum(sales))

print(t_region_sum)
```


### Two or more groups

We can group by more than 1 field.

```{r, eval=T, echo=TRUE, message=F }
library(tidyverse)

t <- tibble(
  region = c('a', 'b', 'c', 'a', 'a', 'b'),
  country = c('USA', 'USA', 'DEN', 'USA', 'USA', 'USA'),
  sales = c(0, 10, 5, 3, 2, 1)
)

# Show the sum of dales by region
t_region_sum <- t %>% 
  group_by(country, region) %>% 
  summarise(sum_of_sales = sum(sales))

print(t_region_sum)
```



### Functions

There are a number of useful functions that can be used in the `summarize` function.

```{r, eval=T, echo=TRUE, message=F }
library(tidyverse)

t <- tibble(
  region = c('a', 'b', 'c', 'a', 'a', 'b'),
  sales = c(0, 10, 5, 3, 2, 1)
)

# Show the sum of dales by region
t_region_stats <- t %>% 
  group_by(region) %>% 
  summarise(n_of_rows = n(),
            sum_of_sales = sum(sales),
            median_sales = median(sales),
            max_sales = max(sales),
            min_sales = min(sales),
            sd_sales = sd(sales),
            n_of_distinct_sales = n_distinct(sales))

print(t_region_stats)
```


### ungroup

Sometimes we want to remove a group from a tibble. You can call `ungroup()`. In this course, we generally create a new tibble for each summary. But, in the real world, you sometimes want to avoid duplicating tibbles when working with large datasets.

```{r, eval=T, echo=TRUE, message=F }
library(tidyverse)

t <- tibble(
  region = c('a', 'b', 'c', 'a', 'a', 'b'),
  sales = c(0, 10, 5, 3, 2, 1)
)

# Group and then ungroup
t <- t %>% 
  group_by(region) %>% 
  ungroup()

print(t)
```

## pivot

We often want to transition data from a wide format to tall, and vice versa.

### wider

Going wide means we take multiple rows and turn them into a single row with multiple columns.

The two key arguments are:

- `names_from` is a field, each value will turn into a new column
- `values_from` is a field, each value will turn into the data for a new column

```{r, eval=T, echo=TRUE }
library(tidyverse)

t_long <- tibble(
  state = c('CA', 'CA', 'CA', 'TX', 'TX', 'TX'),
  year = c(2000, 2001, 2002, 2000, 2001, 2002),
  sales = c(1, 2, 3, 11, 12, 13)
)

t_wide <- t_long %>% 
  pivot_wider(names_from = year, values_from = sales )

print(t_wide)
```


### longer

Going long/tall means we go from a single row with multiple columns and into multiple rows with a single column.

The three key arguments are:

- `cols` are the fields/columns we should remove
  - Name individual items in a vector, i.e. `c(field1, field2)`
  - Use anything that works in `dplyr::select`
    - `field1:field10` selects all of the columns from one field to another. Very good for year columns
    - `starts_with('value')` picks all fields with the beginning value.
- `names_from` is a "string" to call the new column for the names
- `values_from` is a "string" to call the new column for the values

```{r, eval=T, echo=TRUE }
library(tidyverse)

t_wide <- tibble(
  year = c(2000, 2001, 2002),
  ca = c(1, 2, 3),
  tx = c(11, 12, 13)
)

t_long <- t_wide %>% 
  pivot_longer(cols = c(ca, tx), names_to = 'state', values_to = 'sales' )

print(t_long)
```

## Joins

Similar to SQL, we will use joins in dplyr. Note that this uses very weird 
syntax. There are multiple ways to specify the join, but try to always use
the `by = c("field_1" = "field_2")` version. This is the most 
consistently useful approach.

Here is a more thorough guide: (Let's talk about joins)[https://cghlewis.com/blog/joins/]

```{r, eval=FALSE, echo=TRUE }

# Create sample tibbles.
checks <- tibble(
  check_id = c(1, 2, 3, 4, 5),
  desc = c("Cheese", "Bread", "Milk", "Mustard", "Milk"),
  vendor_fk = c(2, 2, 2, 3, 5),
  amount = c(10, NA, 30, 40, 50)
)

vendors <- tibble(
  vendor_pk = c(2, 3, 4, 9),
  desc = c("Bob", "Sarah", "Tim", "Andrew"),
)

# Create a left, right, and inner join for checks and vendors.
# Note the different results
left_join(checks, vendors, by = c("vendor_fk" = "vendor_pk"))
right_join(checks, vendors, by = c("vendor_fk" = "vendor_pk"))
inner_join(checks, vendors, by = c("vendor_fk" = "vendor_pk"))

# Note that we can have issues with NA values.
sum(checks$amount)

# use na.rm = TRUE to remove.
sum(checks$amount, na.rm = TRUE)

```

# Less commonly used dplyr commands

## slice_min/max {.tabset}

We can use slice_max and slice_min to find the rows with the highest and lowest values in a field.

### slice_min when n = 1

`slice_min` pulls the rows with the lowest value in the given field. It respects the current grouping level. When pulling a single value, it is basically identical to `min(field_name)`. 
`slice_max` is the same, but pulls the largest item instead of the smallest.

```{r, eval=T, echo=TRUE, message=F }
library(tidyverse)

t <- tibble(
  region = c('a', 'b', 'c', 'a', 'a', 'b'),
  sales = c(0, 10, 5, 3, 2, 1)
)

# Group and then find smallest value
t <- t %>% 
  group_by(region) %>% 
  slice_min(sales, n = 1) # same as summarize(sales = min(sales))

print(t)
```


### slice_min when n > 1

When given `n > 1`, slice will pull multiple rows.

```{r, eval=T, echo=TRUE, message=F }
library(tidyverse)

t <- tibble(
  region = c('a', 'b', 'c', 'a', 'a', 'b'),
  sales = c(0, 10, 5, 3, 2, 1)
)

# Group and then find the 2 smallest items from each group
t <- t %>% 
  group_by(region) %>% 
  slice_min(sales, n = 2) 

print(t)
```

## rename 

Rename allows you to change the name of a field. This is really handy when dealing with badly-named fields containing spaces.

Note that we use the `backtick` for naming fields with spaces. These are different from `'single quotes'` or `"double quotes"`.

```{r, eval=T, echo=TRUE, message=F }
library(tidyverse)

# Create sample tibbles.
checks <- tibble(
  `Item Name` = c("Cheese", "Bread", "Milk", "Mustard", "Milk"))

checks <- checks %>% 
  rename(item_name = `Item Name`)

print(checks)
```
## relocate

The relocate verb moves columns in a table. It is handy when you are looking for a table to be in a certain order.

# Other concepts

## pipe operator

We can put these all together using the `%\>%` (pipe) operator. This joins
multiple results without needing to re-type stuff.

Note that all dplyr functions use the tibble as the first argument.
Rather than nesting or using intermediate variables, %\>% allows us to
rewrite them.

`x %\>% f(y)` turns into `f(x, y)`

Read left-to-right, top-to-bottom.

You can type this with *Control+Shift+M*

```{r, eval=FALSE, echo=F }
# Application

# We can save each intermediate step.
p <- select(pirates, sex, age, college, beard.length, grogg)
p <- mutate(p, is_male = p$sex == "male")
p <- mutate(p, is_female = p$sex == "female")
p <- arrange(p, p$age)


# Or just use the pipe
p <- tibble(pirates) %>%
  select(sex, age, college, beard.length, grogg) %>%
  mutate(isCCCC = college == "CCCC") %>%
  mutate(is_male = sex == "male") %>%
  mutate(is_female = sex == "female") %>%
  arrange(age)

print(p)
```


## References

The book "R for Data Science" is very good. Below are several key chapters:

- [tibble](https://r4ds.had.co.nz/tibbles.html)
- [Data import](https://r4ds.had.co.nz/data-import.html)
- [Tidy Data](https://r4ds.had.co.nz/tidy-data.html)
- [Relational data](https://r4ds.had.co.nz/relational-data.html)
