---
title: dplyr
output: 
  html_document:
    toc: true
    theme: united
---

dplyr is an important tool for fixing data.   If you need further help,
 see [dplyr tutorial](https://dplyr.tidyverse.org/articles/dplyr.html)

These notes correspond to Datacamp's *Introduction to the tidyverse*.

# Data Visualizations

There are a number of helpful websites with animations of dplyr functions.

- (https://www.andrewheiss.com/blog/2024/04/04/group_by-summarize-ungroup-animations/)
- (https://github.com/nischalshrestha/Unravel)


# Why tibbles?

Data.frames are an older technology, and tibbles are their newer
version. Tibbles do not  automatically change data types. A data.frame guesses 
at times, but a  tibble must be told exactly what to do with incoming data. 
For example, a factor must be manually created. 
Tibbles also are better at printing.

# Converting from data.frame to tibble

If you've used data.frames, you may have used the approach below for subsetting.
The syntax for filtering rows/columns is pretty twitchy. dplyr is easier.

```{r, eval=F, echo=T, message=FALSE }
library(tidyverse)

# Create a data.frame
df <- data.frame(x = c(1, 2, 3, 4, 5), y = c("a", "b", "c", "d", "e"))

# Only include values with x>3
# Note that we use [rows_to_include, columns_to_include]
df[df$x > 3, ]

# Subset, finding a specific column. Return different types.
typeof(df[1])
typeof(df$x)
is.vector(df$x)
class(df[1])

```

# Creating a tibble

You will eventually load tibbles data from a file, but for now I will
usually include them in the file.

```{r, eval=F, echo=TRUE, message=FALSE }
library(tidyverse)

# a tibble created from vectors.
# Note that either they must have the same length,
# or a length of zero (which are automatically expanded)
t <- tibble(
  x = c(1, 2, 3, 4),
  y = c(5,  6,  7,  9),
  z = c("?")
)

# You can subset a tibble and get a tibble as a result
t["x"]
t[c("x", "y")]

# Getting a vector requires double brackets.
t[["x"]]

# You can verify the output by using the class function.
class(t[["x"]])

```

# dplyr commands

## select {.tabset}

Select allows us to choose columns. It helps when we load a large dataset with more columns than we need.

### Columns by name

Pick columns by passing their field names, with each separated by a comma.

```{r, eval=T, echo=TRUE, message=FALSE }
library(tidyverse)

t <- tibble(
  names = c("Bob", "Sarah", "Tim", "John"),
  ages = c(23,  61,  17,  9),
  id = c(1, 2, 3, 4)
)

t_names <- t %>% 
  select(names, ages)

print(t_names)
```


### Remove columns by name

Remove specific columns by placing a `-` in front of each.

```{r, eval=T, echo=TRUE, message=FALSE }
library(tidyverse)

t <- tibble(
  names = c("Bob", "Sarah", "Tim", "John"),
  ages = c(23,  61,  17,  9),
  id = c(1, 2, 3, 4),
  null_field = c(NA, NA, NA, NA)
)

t_no_id_or_null <- t %>% 
  select(-id, -null_field)

print(t_no_id_or_null)
```



### Add columns by name array

You can include a list of columns with `start_column:end_column`.

```{r, eval=T, echo=TRUE, message=FALSE }
library(tidyverse)

t <- tibble(
  extra = 1:4,
  name = c("Bob", "Sarah", "Tim", "John"),
  ages = c(23,  61,  17,  9),
  id = c(1, 2, 3, 4),
  null_field = c(NA, NA, NA, NA)
)

t_name_to_id <- t %>% 
  select(name:id)

print(t_name_to_id)
```


### All numeric columns

You can also use the `where` function pick only fields meeting a condition. 


```{r, eval=T, echo=TRUE, message=FALSE }
library(tidyverse)

t <- tibble(
  extra = 1:4,
  name = c("Bob", "Sarah", "Tim", "John"),
  ages = c(23,  61,  17,  9),
  id = c(1, 2, 3, 4),
  null_field = c(NA, NA, NA, NA)
)

t_numbers <- t %>% 
  select(where(is.numeric))

print(t_numbers)
```


### Helper functions

We can use helper functions to select fields:

- `starts_with()`
- `ends_with()`
- `contains()`
- `matches()`

See the docs for other options.

```{r, eval=T, echo=TRUE, message=FALSE }
library(tidyverse)

t <- tibble(
  id = c(1, 2, 3),
  population_start = c(100, 200, 300),
  population_middle = c(110, 150, 200),
  population_end = c(120, 80, 20),
  not_needed_field = c(NA, "hey", "umn")
)

t_smaller <- t %>% 
  select(starts_with('population'), id)

print(t_smaller)
```


## filter {.tabset}

Removes rows (observations). Uses the symbols:

- `&` as *and*
- `!` as *or*
- `|` as *not*

### Basic example

```{r, eval=T, echo=TRUE, message=FALSE }
library(tidyverse)

t <- tibble(sex = c("man", "woman", "man", "woman", "man"))

t_men_only <- t %>%
  filter(sex == "man")

print(t_men_only)
```

### And

You can combine multiple logical tests with 3 different approaches

```{r, eval=T, echo=TRUE, message=FALSE }
library(tidyverse)
t <- tibble(sex = c("man", "woman", "man", "woman", "man"),
            age = c(10, 20, 30, 40, 50))


# Use one the following:

# Option 1: Use & between each test
t_men_over_30 <- t %>%
  filter(sex == "man" & age > 30)

# Option 2: Separate each test with a comma
t_men_over_30 <- t %>%
  filter(sex == "man",  age > 30)

# Option 3: Use a separate filter for each logical test
t_men_over_30 <- t %>%
  filter(sex == "man") %>%
  filter(age > 30)

print(t_men_over_30)
```

### Or

The `|` symbol is used when we want to satisfy *at least one* of several logical tests.

```{r, eval=T, echo=TRUE, message=FALSE }
library(tidyverse)
t <- tibble(sex = c("man", "woman", "man", "woman", "man"),
            age = c(10, 20, 30, 40, 50))

t_men_or_those_over_30 <- t %>%
  filter(sex == "man" | age > 30)

print(t_men_or_those_over_30)
```


### Not

Use the `!` symbol to test for anything *not true* (or *false*).

```{r, eval=T, echo=TRUE, message=FALSE }
library(tidyverse)
t <- tibble(sex = c("man", "woman", "man", "woman", "man"),
            age = c(10, 20, 30, 40, 50))

t_not_age_30 <- t %>%
  filter(!age == 30)

print(t_not_age_30)
```

### Test for NA values

Test for `NA` values with `is.na()`. You can remove `NA` values  by filtering for `!is.na(column_name)`.

Each `NA` value is unique. So, testing for `column_name == NA` will always be `FALSE`.

```{r, eval=T, echo=TRUE, message=FALSE }
library(tidyverse)
t <- tibble(sex = c("man", "woman", NA))

t_not_na <- t %>%
  filter(!is.na(sex))

print(t_not_na)
```


### Filter by list

We can use the `%in%` command to find values in a vector.

```{r, eval=T, echo=TRUE, message=FALSE }
library(tidyverse)
t <- tibble(tool = c("driver", "saw", "nail", "hammer"))

t2 <- t %>%
  filter(tool %in% c("nail", "hammer"))

print(t2)
```



## mutate {.tabset}

Adds a new column or changes an existing one.

### Change values in a column 

```{r, eval=T, echo=TRUE, message=FALSE }
library(tidyverse)
# population in millions
cities <- tibble(population = c(10, 20, 15, 2, 3))

# Change population from millions to thousands
cities <- cities %>% 
  mutate(population = population * 1000)

print(cities)
```

### Change value to ratio

We can change a value into a percentage (or ratio) by using `sum()` and then dividing a field by that value.

```{r, eval=T, echo=TRUE, message=FALSE }
library(tidyverse)
# population in millions
cities <- tibble(population = c(10, 20, 15, 2, 3))

total_population <- sum(cities$population)

# Change population from millions to thousands
cities <- cities %>% 
  mutate(population = population / total_population)

print(cities)
```

### New Boolean column 

```{r, eval=T, echo=TRUE, message=FALSE }
library(tidyverse)
t <- tibble(sex = c("man", "woman", "man", "woman", "man"))

t_is_man <- t %>% 
  mutate(is_man = sex == "man")

print(t_is_man)
```

### ifelse returning 1/0

We typically use `ifelse()` to create new columns. The first argument will be a *logical test*. If the test is `TRUE`, the function will return its second argument. If the test is `FALSE`, it will return the third argument.

```{r, eval=T, echo=TRUE, message=FALSE }
library(tidyverse)
t <- tibble(sex = c("man", "woman", "man", "woman", "man"))

# Function:
#    ifelse(*logical_test*, *result_if_true*, *result_if_false*)
# 
# Below gives a 1 for males, and a 0 for all others
# 
# Note that we have to end with two parenthesis, one for the ifelse and one for the mutate.
t_with_is_man <- t %>% 
  mutate(is_man01 = ifelse(sex == "man", 1, 0) )

print(t_with_is_man)
```


### ifelse returning string

This example shows `ifelse()` returning a string.

```{r, eval=T, echo=TRUE, message=FALSE }
library(tidyverse)
t <- tibble(sex = c("man", "woman", "man", "woman", "man"))

t_with_m <- t %>% 
  mutate(s = ifelse(sex == "man", "m", "w") )

print(t_with_m)
```


### Multiple new columns

We can add multiple columns with a single mutate.
Below adds two new columns and then modifies the `sex` field.

```{r, eval=T, echo=TRUE, message=FALSE }
library(tidyverse)
t <- tibble(sex = c("Man", "Woman", "MAN", "wOMan", "man"))

t2 <- t %>% 
  mutate(sex_capitalized = str_to_upper(sex),
         sex_lower = str_to_lower(sex),
         sex = str_to_title(sex))

print(t2)
```


## arrange {.tabset}

The `arrange` functions sorts data.

### Ascending  

```{r, eval=T, echo=TRUE, message=FALSE }
library(tidyverse)
t <- tibble(sex = c("man", "woman", "man", "woman", "man"),
            age = c(20, 30, 40, 30, 50))

t_sorted <- t %>% 
  arrange(sex)

print(t_sorted)
```


### Ascending with two or more columns 

```{r, eval=T, echo=TRUE, message=FALSE }
library(tidyverse)
t <- tibble(sex = c("man", "woman", "man", "woman", "man"),
            age = c(20, 30, 40, 30, 50))

t_sorted <- t %>% 
  arrange(sex, age)

print(t_sorted)
```

### Descending order 

Wrap a field with `desc(field_name)` to reverse sort (z-a).

```{r, eval=T, echo=TRUE, message=FALSE }
library(tidyverse)
t <- tibble(sex = c("man", "woman", "man", "woman", "man"),
            age = c(20, 30, 40, 30, 50))

t_sorted <- t %>% 
  arrange(desc(sex), desc(age))

print(t_sorted)
```

# Less commonly used dplyr commands

## rename 

Rename allows you to change the name of a field. This is really handy when dealing with badly-named fields containing spaces.

Note that we use the `backtick` for naming fields with spaces. These are different from `'single quotes'` or `"double quotes"`.

```{r, eval=T, echo=TRUE, message=F }
library(tidyverse)

# Create sample tibbles.
checks <- tibble(
  `Item Name` = c("Cheese", "Bread", "Milk", "Mustard", "Milk"))

checks <- checks %>% 
  rename(item_name = `Item Name`)

print(checks)
```
## relocate

The relocate verb moves columns in a table. It is handy when you are looking for a table to be in a certain order.



# pipe operator

We can put these all together using the `%\>%` (pipe) operator. This joins
multiple results without needing to re-type stuff.

Note that all dplyr functions use the tibble as the first argument.
Rather than nesting or using intermediate variables, %\>% allows us to
rewrite them.

`x %\>% f(y)` turns into `f(x, y)`

Read left-to-right, top-to-bottom.

You can type this with *Control+Shift+M*

```{r, eval=FALSE, echo=F }
# Application

# We can save each intermediate step.
p <- select(pirates, sex, age, college, beard.length, grogg)
p <- mutate(p, is_male = p$sex == "male")
p <- mutate(p, is_female = p$sex == "female")
p <- arrange(p, p$age)


# Or just use the pipe
p <- tibble(pirates) %>%
  select(sex, age, college, beard.length, grogg) %>%
  mutate(isCCCC = college == "CCCC") %>%
  mutate(is_male = sex == "male") %>%
  mutate(is_female = sex == "female") %>%
  arrange(age)

print(p)
```


## References

The book "R for Data Science" is very good. Below are several key chapters:

- [tibble](https://r4ds.had.co.nz/tibbles.html)
- [Data import](https://r4ds.had.co.nz/data-import.html)
- [Tidy Data](https://r4ds.had.co.nz/tidy-data.html)
- [Relational data](https://r4ds.had.co.nz/relational-data.html)


# Application problem

See [problems on GitHub](https://github.com/profgarrett/course_eda_problems/tree/main/r10-dplyr)