---
title: dplyr
output: 
  html_document:
    toc: true
    theme: united
---

This functions in dplyr allow us to reshape data. Instead of changing individual columns, we can instead compress data down or move from wide to tall format. 

See [dplyr tutorial](https://dplyr.tidyverse.org/articles/dplyr.html) for more help.


## group_by and summarize {.tabset}

These two command are similar to SQL's `GROUP BY` or Excel's *Pivot Table*.

The `group_by` command is used first. It doesn't change the data, but instead creates a note on the tibble. 

The `summarize` command is used second. It reads the note on the tibble set by `group_by` to find which fields to include in the results. It then performs some type of aggregation (such as sum or count). The results are a new tibble.


### Summarize without group_by

`summarize()` is similar to mutate, in that both create a new column. But, each new field must have an aggregation function on it, such as `sum()`.

```{r, eval=T, echo=TRUE, message=F }
library(tidyverse)

t <- tibble(
  region = c('a', 'b', 'c', 'a', 'a', 'b'),
  sales = c(0, 10, 5, 3, 2, 1)
)

# Show the sum of dales by region
t_sum <- t %>% 
  summarise(sum_of_sales = sum(sales))

print(t_sum)
```

### Summarize with group_by

While `summarize` is helpful, we normally want to break the results down by some of the fields.

Start by using `group_by`, which includes region in our results. Then use `summarize()` to show which field we want to aggregate.

`summarize()` is similar to mutate, in that both create a new column.

```{r, eval=T, echo=TRUE, message=F }
library(tidyverse)

t <- tibble(
  region = c('a', 'b', 'c', 'a', 'a', 'b'),
  sales = c(0, 10, 5, 3, 2, 1)
)

# Show the sum of dales by region
t_region_sum <- t %>% 
  group_by(region) %>% 
  summarise(sum_of_sales = sum(sales))

print(t_region_sum)
```


### Two or more groups

We can group by more than 1 field.

```{r, eval=T, echo=TRUE, message=F }
library(tidyverse)

t <- tibble(
  region = c('a', 'b', 'c', 'a', 'a', 'b'),
  country = c('USA', 'USA', 'DEN', 'USA', 'USA', 'USA'),
  sales = c(0, 10, 5, 3, 2, 1)
)

# Show the sum of dales by region
t_region_sum <- t %>% 
  group_by(country, region) %>% 
  summarise(sum_of_sales = sum(sales))

print(t_region_sum)
```



### Functions

There are a number of useful functions that can be used in the `summarize` function.

```{r, eval=T, echo=TRUE, message=F }
library(tidyverse)

t <- tibble(
  region = c('a', 'b', 'c', 'a', 'a', 'b'),
  sales = c(0, 10, 5, 3, 2, 1)
)

# Show the sum of dales by region
t_region_stats <- t %>% 
  group_by(region) %>% 
  summarise(n_of_rows = n(),
            sum_of_sales = sum(sales),
            median_sales = median(sales),
            max_sales = max(sales),
            min_sales = min(sales),
            sd_sales = sd(sales),
            n_of_distinct_sales = n_distinct(sales))

print(t_region_stats)
```


### ungroup

Sometimes we want to remove a group from a tibble. You can call `ungroup()`. In this course, we generally create a new tibble for each summary. But, in the real world, you sometimes want to avoid duplicating tibbles when working with large datasets.

```{r, eval=T, echo=TRUE, message=F }
library(tidyverse)

t <- tibble(
  region = c('a', 'b', 'c', 'a', 'a', 'b'),
  sales = c(0, 10, 5, 3, 2, 1)
)

# Group and then ungroup
t <- t %>% 
  group_by(region) %>% 
  ungroup()

print(t)
```

## pivot

We often want to transition data from a wide format to tall, and vice versa.

### wider

Going wide means we take multiple rows and turn them into a single row with multiple columns.

The two key arguments are:

- `names_from` is a field, each value will turn into a new column
- `values_from` is a field, each value will turn into the data for a new column

```{r, eval=T, echo=TRUE }
library(tidyverse)

t_long <- tibble(
  state = c('CA', 'CA', 'CA', 'TX', 'TX', 'TX'),
  year = c(2000, 2001, 2002, 2000, 2001, 2002),
  sales = c(1, 2, 3, 11, 12, 13)
)

t_wide <- t_long %>% 
  pivot_wider(names_from = year, values_from = sales )

print(t_wide)
```


### longer

Going long/tall means we go from a single row with multiple columns and into multiple rows with a single column.

The three key arguments are:

- `cols` are the fields/columns we should remove
  - Name individual items in a vector, i.e. `c(field1, field2)`
  - Use anything that works in `dplyr::select`
    - `field1:field10` selects all of the columns from one field to another. Very good for year columns
    - `starts_with('value')` picks all fields with the beginning value.
- `names_from` is a "string" to call the new column for the names
- `values_from` is a "string" to call the new column for the values

```{r, eval=T, echo=TRUE }
library(tidyverse)

t_wide <- tibble(
  year = c(2000, 2001, 2002),
  ca = c(1, 2, 3),
  tx = c(11, 12, 13)
)

t_long <- t_wide %>% 
  pivot_longer(cols = c(ca, tx), names_to = 'state', values_to = 'sales' )

print(t_long)
```

## Joins

Similar to SQL, we will use joins in dplyr. Note that this uses very weird 
syntax. There are multiple ways to specify the join, but try to always use
the `by = c("field_1" = "field_2")` version. This is the most 
consistently useful approach.

Here is a more thorough guide: (Let's talk about joins)[https://cghlewis.com/blog/joins/]

```{r, eval=FALSE, echo=TRUE }

# Create sample tibbles.
checks <- tibble(
  check_id = c(1, 2, 3, 4, 5),
  desc = c("Cheese", "Bread", "Milk", "Mustard", "Milk"),
  vendor_fk = c(2, 2, 2, 3, 5),
  amount = c(10, NA, 30, 40, 50)
)

vendors <- tibble(
  vendor_pk = c(2, 3, 4, 9),
  desc = c("Bob", "Sarah", "Tim", "Andrew"),
)

# Create a left, right, and inner join for checks and vendors.
# Note the different results
left_join(checks, vendors, by = c("vendor_fk" = "vendor_pk"))
right_join(checks, vendors, by = c("vendor_fk" = "vendor_pk"))
inner_join(checks, vendors, by = c("vendor_fk" = "vendor_pk"))

# Note that we can have issues with NA values.
sum(checks$amount)

# use na.rm = TRUE to remove.
sum(checks$amount, na.rm = TRUE)

```

# Less commonly used dplyr commands

## slice_min/max {.tabset}

We can use slice_max and slice_min to find the rows with the highest and lowest values in a field.

### slice_min when n = 1

`slice_min` pulls the rows with the lowest value in the given field. It respects the current grouping level. When pulling a single value, it is basically identical to `min(field_name)`. 
`slice_max` is the same, but pulls the largest item instead of the smallest.

```{r, eval=T, echo=TRUE, message=F }
library(tidyverse)

t <- tibble(
  region = c('a', 'b', 'c', 'a', 'a', 'b'),
  sales = c(0, 10, 5, 3, 2, 1)
)

# Group and then find smallest value
t <- t %>% 
  group_by(region) %>% 
  slice_min(sales, n = 1) # same as summarize(sales = min(sales))

print(t)
```


### slice_min when n > 1

When given `n > 1`, slice will pull multiple rows.

```{r, eval=T, echo=TRUE, message=F }
library(tidyverse)

t <- tibble(
  region = c('a', 'b', 'c', 'a', 'a', 'b'),
  sales = c(0, 10, 5, 3, 2, 1)
)

# Group and then find the 2 smallest items from each group
t <- t %>% 
  group_by(region) %>% 
  slice_min(sales, n = 2) 

print(t)
```

## rename 

Rename allows you to change the name of a field. This is really handy when dealing with badly-named fields containing spaces.

Note that we use the `backtick` for naming fields with spaces. These are different from `'single quotes'` or `"double quotes"`.

```{r, eval=T, echo=TRUE, message=F }
library(tidyverse)

# Create sample tibbles.
checks <- tibble(
  `Item Name` = c("Cheese", "Bread", "Milk", "Mustard", "Milk"))

checks <- checks %>% 
  rename(item_name = `Item Name`)

print(checks)
```

## relocate

The relocate verb moves columns in a table. It is handy when you are looking for a table to be in a certain order.



## References

The book "R for Data Science" is very good. Below are several key chapters:

- [tibble](https://r4ds.had.co.nz/tibbles.html)
- [Data import](https://r4ds.had.co.nz/data-import.html)
- [Tidy Data](https://r4ds.had.co.nz/tidy-data.html)
- [Relational data](https://r4ds.had.co.nz/relational-data.html)
