---
title: dplyr2 - shaping
output: 
  html_document:
    toc: true
    theme: united
---

These functions in dplyr allow us to reshape data.  See [dplyr tutorial](https://dplyr.tidyverse.org/articles/dplyr.html) for more help.


## group_by and summarize {.tabset}

These two command are similar to SQL's `GROUP BY` or Excel's *Pivot Table*.

First, the `group_by` command adds a note on the tibble. It does not change any of the data.

Second, the `summarize` command combines rows. It reads the note on the tibble set by `group_by`. That field will be used as the labels in my smaller tibble. `summarize` then performs some type of aggregation (such as sum or count). 


### Summarize without group_by

Both `summarize` and `mutate` create a new column. But, `summarize` requires that each new field have function, such as `sum` or `sd`. `summarize` also reduces the number of rows, one for each unique value in the `group_by` note.

```{r eval=T, echo=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)

t <- tibble(
  region = c('a', 'b', 'c', 'a', 'a', 'b'),
  sales = c(0, 10, 5, 3, 2, 1)
)

# Show the sum of dales by region
t_sum <- t %>% 
  summarise(sum_of_sales = sum(sales))

print(t_sum)
```

### Summarize with group_by

While `summarize` is helpful, we normally want to the results grouped by a field.

Start by using `group_by`, which includes region in our results. Then use `summarize` to show which field we want to aggregate.


```{r, eval=T, echo=TRUE, message=F }
library(tidyverse)

t <- tibble(
  region = c('a', 'b', 'c', 'a', 'a', 'b'),
  sales = c(0, 10, 5, 3, 2, 1)
)

# Show the sum of sales by region
t_region_sum <- t %>% 
  group_by(region) %>% 
  summarise(sum_of_sales = sum(sales))

print(t_region_sum)
```


### Two or more groups

We can group by more than one field.

```{r, eval=T, echo=TRUE, message=F }
library(tidyverse)

t <- tibble(
  region = c('a', 'b', 'c', 'a', 'a', 'b'),
  country = c('USA', 'USA', 'DEN', 'USA', 'USA', 'USA'),
  sales = c(0, 10, 5, 3, 2, 1)
)

# Show the sum of dales by region
t_region_sum <- t %>% 
  group_by(country, region) %>% 
  summarise(sum_of_sales = sum(sales))

print(t_region_sum)
```



### Functions

There are a number of useful functions.

```{r, eval=T, echo=TRUE, message=F }
library(tidyverse)

t <- tibble(
  region = c('a', 'b', 'c', 'a', 'a', 'b'),
  sales = c(0, 10, 5, 3, 2, 1)
)

# Show the sum of sales by region
t_region_stats <- t %>% 
  group_by(region) %>% 
  summarise(n_of_rows = n(),
            sum_of_sales = sum(sales),
            median_sales = median(sales),
            mean_sales = mean(sales),
            max_sales = max(sales),
            min_sales = min(sales),
            sd_sales = sd(sales),
            n_of_distinct_sales = n_distinct(sales))

print(t_region_stats)
```



### Function NA problem

Functions will have issues if there is a NA.
Use an argument to remove all NAs

```{r, eval=T, echo=TRUE, message=F }
library(tidyverse)

t <- tibble(
  region = c('a', 'a', 'b', 'b'),
  sales = c(0, 10, 5, NA)
)

# Show the sum of dales by region
t_region_stats <- t %>% 
  group_by(region) %>% 
  summarise(n_of_rows = n(),
            sum_of_sales = sum(sales, na.rm = TRUE),
            median_sales = median(sales, na.rm = TRUE),
            mean_sales = mean(sales, na.rm = TRUE),
            max_sales = max(sales, na.rm = TRUE),
            min_sales = min(sales, na.rm = TRUE),
            sd_sales = sd(sales, na.rm = TRUE),
            n_of_distinct_sales = n_distinct(sales, na.rm = TRUE))

print(t_region_stats)
```



### ungroup

Sometimes we want to remove a group from a tibble. You can call `ungroup`. In this course, we generally create a new tibble for each summary. But, in the real world, you sometimes want to avoid duplicating tibbles when working with large datasets.

```{r, eval=T, echo=TRUE, message=F }
library(tidyverse)

t <- tibble(
  region = c('a', 'b', 'c', 'a', 'a', 'b'),
  sales = c(0, 10, 5, 3, 2, 1)
)

# Group and then ungroup
t <- t %>% 
  group_by(region) %>% 
  ungroup()

print(t)
```

## Joins {.tabset}

Joins are used to combine multiple tibbles. They are very similar to joins in SQL.

Here is a more thorough guide: [Let's talk about joins](https://cghlewis.com/blog/joins/)

### inner_join

Inner join will return all rows that exist in both tables.

```{r echo=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)

# Create sample tibbles.
checks <- tibble(
  vendor_pk = c(2, 2, 2, 3, 5),
  amount = c(10, 15, 30, 40, 50)
)

vendors <- tibble(
  vendor_pk = c(2, 3, 4, 9),
  vendor_name = c("Bob", "Sarah", "Tim", "Andrew"),
)

# Note that we get the same results if we start with either vendor or checks
joined_table <- checks %>% 
  inner_join(vendors, by = "vendor_pk")

print(joined_table)
```


### left_join

Left (outer) join will return all rows from one table, and matching rows in the second table. 

Order matters! When using a left join, whatever table you start with will have all rows returned. Only matching rows from the right will be returned.


```{r echo=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)

# Create sample tibbles.
checks <- tibble(
  vendor_pk = c(2, 2, 2, 3, 5),
  amount = c(10, 15, 30, 40, 50)
)

vendors <- tibble(
  vendor_pk = c(2, 3, 4, 9),
  vendor_name = c("Bob", "Sarah", "Tim", "Andrew"),
)

# Note that we get *different* results if we start with either vendor or checks
# Because we start with checks, all checks are included (and only matching vendors)
joined_table <- checks %>% 
  left_join(vendors, by = "vendor_pk")

print(joined_table)
```

### rename when fields have different names

Rename allows you to change the name of a field. This is really handy when each tibble has a different name for a field.


```{r echo=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)

# Create sample tibbles.
checks <- tibble(
  check_id = c(1, 2, 3, 4),
  check_amount = c(100, 200, 150, 250)
)

invoices <- tibble(
  invoice_key = c(10, 20, 30, 40),
  check_key = c(1, 2, 3, 4)
)

# Start with checks, rename its field to match the 
# name in the other table, and then join using the new name.
joined_data <- checks %>% 
  rename(check_key = check_id) %>% 
  inner_join(invoices, by = 'check_key')

print(joined_data)
```

# Less common commands

## slice_min/max {.tabset}

We can use slice_max and slice_min to find the rows with the highest and lowest values in a field.

### slice_min when n = 1

`slice_min` pulls the rows with the lowest value in the given field. It respects the current grouping level. When pulling a single value, it is basically identical to `min(field_name)`. 
`slice_max` is the same, but pulls the largest item instead of the smallest.

```{r eval=T, echo=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)

t <- tibble(
  region = c('a', 'b', 'c', 'a', 'a', 'b'),
  sales = c(0, 10, 5, 3, 2, 1)
)

# Group and then find smallest value
t <- t %>% 
  group_by(region) %>% 
  slice_min(sales, n = 1) # same as summarize(sales = min(sales))

print(t)
```


### slice_min when n > 1

When given `n > 1`, slice will pull multiple rows.

```{r eval=T, echo=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)

t <- tibble(
  region = c('a', 'b', 'c', 'a', 'a', 'b'),
  sales = c(0, 10, 5, 3, 2, 1)
)

# Group and then find the 2 smallest items from each group
t <- t %>% 
  group_by(region) %>% 
  slice_min(sales, n = 2) 

print(t)
```

## relocate

The relocate verb moves columns in a table. It is handy when you are looking for a table to be in a certain order.


# Application problem

See [problems on GitHub](https://github.com/profgarrett/course_eda_problems/tree/main/r11-dplyr)
