---
title: dplyr3 - reshaping
output: 
  html_document:
    toc: true
    theme: united
---

This functions allow us to reshape data. Instead of changing individual columns, we can instead move from wide to tall format. 

## pivot

We often want to transition data from a wide format to tall, and vice versa.

This was moved from `dplyr` to the `tidyr` package. But, including tidyverse will load all of the appropriate libraries.

### wider

Going wide means we take multiple rows and turn them into a single row with multiple columns.

The two key arguments are:

- `names_from` is a field, each value will turn into a new column
- `values_from` is a field, each value will turn into the data for a new column

```{r eval=T, echo=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)

t_long <- tibble(
  state = c('CA', 'CA', 'CA', 'TX', 'TX', 'TX'),
  year = c(2000, 2001, 2002, 2000, 2001, 2002),
  sales = c(1, 2, 3, 11, 12, 13)
)

t_wide <- t_long %>% 
  pivot_wider(names_from = year, values_from = sales )

print(t_wide)
```


### longer

Going long/tall means we go from a single row with multiple columns and into multiple rows with a single column.

The three key arguments are:

- `cols` are the fields/columns we should remove
  - Name individual items in a vector, i.e. `c(field1, field2)`
  - Use anything that works in `dplyr::select`
    - `field1:field10` selects all of the columns from one field to another. Very good for year columns
    - `starts_with('value')` picks all fields with the beginning value.
- `names_from` is a "string" to call the new column for the names
- `values_from` is a "string" to call the new column for the values

```{r eval=T, echo=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)

t_wide <- tibble(
  year = c(2000, 2001, 2002),
  ca = c(1, 2, 3),
  tx = c(11, 12, 13)
)

t_long <- t_wide %>% 
  pivot_longer(cols = c(ca, tx), names_to = 'state', values_to = 'sales' )

print(t_long)
```

