---
title: Reading in datafiles
output: 
  html_document:
    toc: true
    theme: united
---

How do we get data into R?


# read_csv {.tabset}

There are a collection of read functions, which all use similar parameters.

## Basic example

This will read a text file that looks like this:
```text
id,name
1,"Bob"
2,"Sam"
```

We only need to pass the name of the file.

```{r include=FALSE}
cat('id,name\n1,"Bob"\n2,"Sam"\n', file = 'r15-read.txt')
```


```{r echo=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)

t <- read_csv("r15-read.txt")
print(t)
```


## Skip starting lines

This datafile has some starting lines that we need to skip

```text
Here is a datafile

id,name
1,"Bob"
2,"Sam"
```

Tell read_csv to skip the first 2 lines.

```{r include=FALSE}
cat('\n\nid,name\n1,"Bob"\n2,"Sam"\n', file = 'r15-read.txt')
```


```{r echo=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)

t <- read_csv("r15-read.txt", skip = 2)
print(t)
```



## Fix missing column names

Some datafiles are missing column names.

```text
1,"Bob"
2,"Sam"
```

Tell read_csv the names for columns

```{r include=FALSE}
cat('1,"Bob"\n2,"Sam"\n', file = 'r15-read.txt')
```


```{r echo=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)

t <- read_csv("r15-read.txt", col_names = c('id', 'name'))
print(t)
```


## Fix NA values 

Some data files have missing values. This can take different forms, such as blank strings, spaces, "magic" numbers, or "NA"

```text
id,name
1, 
NA,"Sam"
-1,"Joe"
```

Tell read_csv which values should be transformed into `NA`.

```{r include=FALSE}
cat('id,name\n1,""\nNA,"Sam"\n-1,"Joe"', file = 'r15-read.txt')
```


```{r echo=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)

t <- read_csv("r15-read.txt", na = c('', ' ', 'NA', '-1'))
print(t)
```

# read_xlsx {.tabset}

We can use `read_xlsx` to load *Excel* files.

## Basic example

We typically need to provide the name of the sheet.

```{r echo=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)
library(readxl)

t <- read_xlsx("r15-read.xlsx", sheet = 'Sheet of Data')
print(t)
```


## Complex example

We can use similar options as `read_csv`


```{r echo=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)
library(readxl)

t <- read_xlsx("r15-read.xlsx", 
               sheet = 'Weird Data Sheet',
               na = c('', ' ', 'NA', '-1'),
               skip = 2)
print(t)
```



# Janitor

Janitor is a helpful library that cleans up badly-formatted column titles. I generally use the function that fixes titles.  

Treat it like `dplyr`, using the pipe symbol as part of your clean-up code.

```{r echo=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)

bad_tibble <- tibble(
  `Title with spaces and CAPITALIZATION` = c(1, 2, 3, 4)
)

good_tibble <- bad_tibble %>% 
  janitor::clean_names()

print(good_tibble)
```

