---
title: ggplot 2 - making a pretty plot
output: 
  html_document:
    toc: true
    theme: united
---

# Introduction

```{r, include=FALSE}

# Add libraries
library(tidyverse)

```


Below are a number of useful functions to improve the attractiveness of your plots.

## Geoms for annotations

### geom_vline / geom_hline

Draw a line or box on the chart.

```{r, eval=T, echo=TRUE }
ggplot(data = mpg) +
  geom_point(mapping = aes(x = cty, y = hwy)) +
  geom_vline(xintercept = 20) +
  geom_hline(yintercept = 30) +
  geom_rect(xmin = 20, xmax = 25, ymin = 25, ymax = 30,
            alpha = 0.005,
            fill = 'green')

```


### geom_label_repel

Add labels with geom_label_repel. There is also a geom_text, but it will
plot labels on top of the data points.

```{r, eval=T, echo=TRUE }
library(ggrepel)

mpg_2seater <- filter(mpg, class == '2seater')

ggplot(data = mpg_2seater) +
  geom_point(mapping = aes(x = cty, y = hwy)) +
  geom_label_repel(mapping = aes(x = cty, y = hwy, label = model))

```




## Labels

We can add a variety of labels to a plot.

```{r, eval=T, echo=TRUE }

ggplot(data = mpg) +
  geom_bar(mapping = aes(y = manufacturer)) +
  labs(title = 'Main title',
       subtitle = 'Subtitle title',
       caption = 'Caption at bottom of chart') +
  xlab('Label for x axis') +
  ylab('Label for y axis')

```


## Theme

There are some nice options for themes. Some include: 

- theme_bw
- theme_classic
- theme_minimal
- theme_void

```{r, eval=T, echo=TRUE }
ggplot(data = mpg) +
  geom_point(mapping = aes(x = cty, y = hwy)) +
  theme_void()
```

## Limits

Set a max/min for an axis.

```{r, eval=T, echo=TRUE }

ggplot(data = mpg) +
  geom_point(mapping = aes(x = cty, y = hwy)) +
  xlim(0, 20) +
  ylim(0, 20) 
```

## Scales

We can customize the axis scales.

You need to match the type of scale to your datatype. Is the data 
continuous (ie., a number) or discrete (generally text)?

### Discrete Scale

A discrete scale handles a vector of text values.
Set custom `labels` using a vector.

```{r, eval=T, echo=TRUE }

ggplot(data = mpg) +
  geom_point(mapping = aes(y = class, x = hwy)) +
  scale_y_discrete(
                  labels = c('2 Seater', 'Compact Car', 'Midsize', 'Minivan', 
                              'Pickup', 'Sub-compact', 'SUV'),
                  name = 'Car Classification')
  
```

### Continuous Scale

A continuous scale is for a series of numbers.

We can set custom breaks, as well as the min/max.

```{r, eval=T, echo=TRUE }

ggplot(data = mpg) +
  geom_point(mapping = aes(y = hwy, x = hwy)) +
  scale_x_continuous(n.breaks = 5, limits = c(20, 30)) +
  scale_y_continuous(breaks = c(15, 20, 25))
```


### Log  Scale

A log scale helps us see data that grows at an exponential level.

```{r, eval=T, echo=TRUE }

ggplot(data = mpg) +
  geom_point(mapping = aes(y = hwy, x = hwy)) +
  scale_x_log10()
```


### Date Scale

Dates/datetimes are continuous values, but don't use a continuous scale. Use `scale_x_date` and `scale_x_datetime` for additional options.


Our main options are:

- *Labels*
   - `labels = scales::label_date("format string")`
     - Find `"format string"` options by using F1 on label_date, and go to its format section, and click on `strptime()`. Scroll down for a list of options.
     - Examples (note lower versus upper case!)
       - `"%Y-%m-%d"` shows as `'2023-01-09'`
       - `"%H:%M:%S"` shows as `'02:00:00'`
- *Breaks* 
   - `date_breaks = "number periods"`
     - `"number period"` is a combination of a number and a period (such as hour, minute, year, etc...)
     - Examples (use lowercase!): 
       - `1 month` 
       - `3 hours`
- *Limits*
  - `limits = c(start_date, end_end)`
  - Create a date or datetime with lubridate
  - Examples:
    - `limits = c( ymd('2023-01-01'), ymd('2023-01-30')`)`
    - `limits = c( ymd_hm('2023-01-01 06:00am'), ymd_hm('2023-01-01 06:00pm')`


See [ggplot's label_date](https://scales.r-lib.org/reference/label_date.html) for help on the scale.

See [lubridate](https://lubridate.tidyverse.org/) for help on dealing with dates.



```{r, eval=T, echo=TRUE }
library(lubridate)

date_tibble <- tibble(
  open = c(ymd_hm('2023-01-01 8:00am'), 
           ymd_hm('2023-01-02 9:00am'), 
           ymd_hm('2023-01-09 3:30pm'), 
           ymd_hm('2023-01-25 5:45pm'))
)

ggplot(data = date_tibble) +
  geom_point(mapping = aes(y = open, x = (open))) +
  scale_y_datetime(labels = scales::label_date("%Y-%m-%d"), 
                   date_breaks = '1 week',
                   limits = c(
                     ymd_hm('2023-01-01 00:00'), 
                     ymd_hm('2023-02-15 00:00'))
                   ) +
  scale_x_datetime(labels = scales::label_time("%H:%M:%S"), 
                   date_breaks = '100 hours')
  
```


### Label formatting

Breaks also may need formatting to fix labels. 

- label_percent(): show as 30%
  - Use `accuracy = 0.1` to round to 10%, or `0.01` to round to 1%.
- label_dollar(): show with a leading $
- label_comma(): insert a comma and avoid scientific notation

```{r, eval=T, echo=TRUE }

ggplot(data = mpg) +
  geom_point(mapping = aes(y = cty, x = hwy)) +
  scale_x_continuous(labels = scales::label_dollar()) +
  scale_y_continuous(labels = scales::label_percent(accuracy = 1))
  
```