---
title: ggplot 3 - Create maps
output: 
  html_document:
    toc: true
    theme: united
---

# Introduction

We can add GIS data using OpenStreetMap

A simple tutorial is here: https://ggplot2tutor.com/tutorials/streetmaps

This is a good additional step, which also talks about points
and some font options.
https://joshuamccrain.com/tutorials/maps/streets_tutorial.html

A deep tutorial is available here: https://taraskaduk.com/posts/2021-01-18-print-street-maps/

Another good link is: https://wiki.openstreetmap.org/wiki/Map_Features
This gives a list of features you can find.

We'll be using some of their suggestions for the code below.

```{r, ggplot3-setup, eval=FALSE, echo=TRUE}

knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(maps)
library(sf)
library(ggrepel)

# Used to setup environment
library(usethis)

# Used for large downloads
# https://docs.ropensci.org/osmextract/
library(osmextract)

```



## OpenStreetMap with osmextract

We can add GIS data using OpenStreetMap using the osmextract library.

This library makes it easier to do larger maps, dealing with the 
time-out problems in the prior library.

See https://docs.ropensci.org/osmextract/ for instructions.

A simple tutorial is here: https://ggplot2tutor.com/tutorials/streetmaps

This is a good additional step, which also talks about points
and some font options.
https://joshuamccrain.com/tutorials/maps/streets_tutorial.html

A deep tutorial is available here: https://taraskaduk.com/posts/2021-01-18-print-street-maps/

Another good link is: https://wiki.openstreetmap.org/wiki/Map_Features
This gives a list of features you can find.

We'll be using some of their suggestions for the code below.


## Setup environment

It's important to setup a default location for all downloaded files. Otherwise,
R will download them every time you load the program. 

```{r, ggplot3-setupenv, eval=FALSE, echo=TRUE}

# See if the download folder is properly setup.
if( "" == Sys.getenv('OSMEXT_DOWNLOAD_DIRECTORY')) {
  print('STOP! You need to fix the osmext download folder! Read the comments')
  
  # After this runs, run the below code to find the correct working directory.
  getwd()
  
  # Copy the path from getwd(), as you'll use it in the next step.
  
  # Run this code. It will create a .Renviron file for this project.
  # The file will open up, and have no content.
  usethis::edit_r_environ('project')
  
  # Then, add the OSMEXT_DOWNLOAD_DIRECTORY option below in your blank file.
  # Paste the working directory path you got from getwd()
  # 
  # It should look like the below (ignoring the comments, and with your path)
  # 
  # ==== start of .Renviron file ============
  # OSMEXT_DOWNLOAD_DIRECTORY="C:/Users/garrettn/project_map"
  #
  # ====  end of .Renviron file =======
  # 
  # Then, restart R and run this block of code again. Running 
  # Sys.getenv('OSMEXT_DOWNLOAD_DIRECTORY') should return the path for your project.
  # R will now save the map files in the project folder.
} else {
  print(paste('Success! Map data will be saved in', getwd()))
}

```

## Download a map

The below downloads data for WV. You'll see that it downloads the entire state,
which is a lot of data.

It also takes a long time to plot our map. The more points you have,
the longer it takes to load the data.

```{r, ggplot3-download, eval=FALSE, echo=TRUE}

# Download WV Data with oe_get
# We go ahead and grab 3 datasets of different types.

osm_lines = osmextract::oe_get("West Virginia, USA", 
                   # layer = "lines", this is the default option
                   stringsAsFactors = FALSE, 
                   quiet = TRUE)

osm_points = osmextract::oe_get("West Virginia, USA", 
                    layer = "points", 
                    stringsAsFactors = FALSE, 
                    quiet = TRUE)

osm_poly = osmextract::oe_get("West Virginia, USA", 
                    layer = "multipolygons", 
                    stringsAsFactors = FALSE, 
                    quiet = TRUE)


# Need to filter down the data.
# Pick some items.

# Take a look at the various items in highway.
tibble(osm_lines) %>% 
  group_by(highway) %>% 
  summarise(count = n())

# Road types.
minor_road_types <- c("secondary", "secondary_link",
           "service", 
           "residential", "tertiary_link", "road")

major_road_types <- c("motorway", "primary", "primary_link", "motorway_link", "trunk" )

wv_major_roads <- filter(osm_lines, highway %in% major_road_types)
wv_minor_roads <- filter(osm_lines, highway %in% minor_road_types & !is.na(name))

# Grab some water elements
wv_water <- tibble(osm_poly) %>% 
            filter(natural == 'water', !is.na(name))

# What admin elements do we have?
unique(osm_poly$admin_level)

# Grab state boundaries
wv_state <- tibble(osm_poly) %>% 
            filter(!is.na(admin_level), admin_level == 4)

ggplot() + 
  geom_sf(data = wv_water$geometry, 
          color = 'steelblue',
          alpha = 1) +
  geom_sf(data = wv_major_roads, 
          color = 'gray',
          alpha = 1) +
  geom_sf(data = wv_state$geometry, 
          fill = NA,
          color = 'black',
          size = 10,
          alpha = 1) +
  coord_sf(xlim = c(-83, -77),
           ylim = c(37, 41),
           expand = FALSE)



```



## Cropping

It's useful to have different ways of cropping the data. The first is st_crop,
which uses x/y min/max values. Note that since WV has a negative X, that 
you should use the numerically smaller / larger values, and not look at the axis.

This is different from coord_sf, which takes a ylim, xlim, and expand = FALSE.
That just restricts the viewport, but still plots all of the data. If we have 
a big dataset, it'll still take a really long time to plot.

```{r, ggplot3-crop, eval=FALSE, echo=TRUE}

# This reduces datapoints in our dataset
cropped_hws <- st_crop(wv_major_roads, 
                         xmin = -82, xmax = -79,
                         ymin = 37.5, ymax = 40)


# coord_sf changes the viewpoint.
# The data is still plotted.


ggplot() +
  geom_sf(data = wv_state$geometry, 
          color = 'black',
          size = 8,
          alpha = 1) +
  geom_sf(data = cropped_hws, 
          color = 'gray',
          alpha = 1) +
  coord_sf(xlim = c(-82.5, -78.5),
           ylim = c(37, 40.5),
           expand = FALSE) +
  theme_bw()

```

We can also use a geography to crop data. Note that this includes points outside
of the boundaries if they start inside of the geometry.

```{r, ggplot3-geo, eval=FALSE, echo=TRUE}

# Grab Charleston boundaries
wv_mon <- tibble(osm_poly) %>% 
            filter(name == 'Charleston')

# This reduces datapoints in our dataset
cropped_hws <- st_crop(wv_major_roads, wv_mon$geometry )

ggplot() +
  geom_sf(data = wv_mon$geometry, 
          color = 'black',
          size = 8,
          alpha = 1) +
  geom_sf(data = cropped_hws, 
          color = 'gray',
          alpha = 1) +
  scale_fill_viridis_c(na.value = "transparent")
  theme_bw()


```

## Exporting

We can export the resulting map with ggsave. This is mostly useful
for when we need to have a very high quality image,
which is hard to see in rStudio.

```{r, ggplot3-export, eval=FALSE, echo=TRUE}

#osm_states = osmextract::oe_get("United States of America",
#                                stringsAsFactors = FALSE,
#                                query = "SELECT * FROM multipolygons WHERE admin_level == 4"
#                                )

osm_lines = osmextract::oe_get("West Virginia, USA", 
                   stringsAsFactors = FALSE, 
                   quiet = TRUE)

# Grab other states   
states <- map_data('state')
states_neighbor <- filter(states, region %in% c('ohio', 'pennsylvania', 
                                                'maryland', 'virginia',
                                                'kentucky'))


wv_major_roads <- filter(osm_lines, highway %in% major_road_types)
wv_minor_roads <- filter(osm_lines, highway %in% minor_road_types & !is.na(name))

p <- ggplot() +
  geom_sf(data = wv_major_roads, 
          color = 'darkgray',
          size = 2,
          alpha = 1) +
  geom_sf(data = wv_minor_roads, 
          color = 'lightgray',
          size = .2,
          alpha = .5) +
  # Add the other states around WV to get their borders,
  # and prevent any bleed-through of items past WV borders.
  geom_polygon(data = states_neighbor,
                 mapping = aes(x = long, y = lat,
                     group = group, fill = region),
                 fill = 'white', 
                 color = "lightgray")  +
  geom_sf(data = wv_state$geometry, 
          color = 'black',
          fill = NA,
          size = 15,
          alpha = 1) +
  coord_sf(xlim = c(-83, -77),
           ylim = c(37, 41),
           expand = FALSE)

    
# Save a very large png
ggsave('output.png', 
       plot = p, 
       width = 7, 
       height = 7, 
       dpi = 320)

```
