---
title: Conditionals in R
output: 
  html_document:
    toc: true
    theme: united
---

# Introduction

Loops are good to know, but are not a critical feature in R. This section corresponds to Datacamp's *Loops* chapter in the *Intermediate R* course.

# While loop

The while feature repeatedly runs some set of code. You can cancel out of a loop with the `break` statement.


```{r eval=TRUE, echo=TRUE}

var_a <- 5

while (var_a > 1) {
  print(var_a)
  var_a <- var_a - 1
  
  if (var_a == 2) {
    print('escape!')
    break;
  }
}

```


# For loop

For loops are easier to use for looping over every element in a vector. You can either loop through each item, or create a new vector of index values (1 to the length of the vector).

The `break` and `next` statements will escape the loop, and skip the remaining code for the current iteration.

```{r eval=TRUE, echo=TRUE}

vector_of_scores <- c(1, 2, 1, 3, 2, 1, 2, 3)

# Approach 1:
#
# Loop through each item
for (score in vector_of_scores) {
  print(score)
}


# Approach 2: Indexes
# 
# Create a vector of indexes
vector_of_indexes <- 1:length(vector_of_scores)
# Loop through each item
for (i in vector_of_indexes) {
  print(paste('Index', i, 'for value', vector_of_scores[i]))
}

```

You can also loop through a matrix with an embedded for loop inside of another for loop. However, we will rarely use this approach in our class.
