---
title: Functions in R
output: 
  html_document:
    toc: true
    theme: united
---

# Introduction

A function is a grouping of code. It helps you organize your code and reduce complexity.


## Calling a function

We can pass values to a function by position and/or name.

``` {r}

my_function <- function(a, b, c) {
  print(a + b + c)
}

# Use position
my_function(1, 2, 3)

# Use names
my_function(a = 1, b = 2, c = 3)

# Use a combination (a will be 1, b will be 2)
my_function(1, 2, c = 3) 

```

We will often rely on default arguments.

``` {r}
my_function <- function(a, increase_by = 10) {
  return(a + increase_by)
}

my_function(1)
my_function(1, increase_by = 1)
```

## Create your own  {.tabset}


### Bad example of duplicated code

Look for examples of copy + pasted code in the same below.
What happens if we want to make an adjustment to all blocks of code?

```{r eval=FALSE, echo=TRUE}
library(tidyverse)

data(mpg)
t_mpg <- mpg


avg_cyl <- mean(t_mpg$cyl)
rnd_avg_cyl <-  round(avg_cyl, 2)
text_avg_cyl <- paste("The average is", rnd_avg_cyl)
print(text_avg_cyl)


avg_hwy <- mean(t_mpg$hwy)
rnd_avg_hwy <-  round(avg_hwy, 2)
text_avg_hwy <- paste("The average highway mpg is ", rnd_avg_hwy)
print(text_avg_hwy)


avg_cty <- mean(t_mpg$cty)
rnd_avg_cty <-  round(avg_cty, 2)
text_avg_cty <- paste("The average city mpg is ", rnd_avg_cty)
print(text_avg_cty)


```


### Fixed duplicated code

Let's break that code out into a separate function.

```{r eval=FALSE, echo=TRUE}
library(tidyverse)

data(mpg)
t_mpg <- mpg

find_avg <- function(text, field) {
  m <- mean(field)
  rnd <-  round(m, 2)
  text <- paste("The average of", text, "is", rnd)
  return(text)
}

print(find_avg("highway mpg", t_mpg$hwy))
print(find_avg("city mpg", t_mpg$cty))


```

### More good examples

Here are some examples of some good functions.

```{r eval=FALSE, echo=TRUE}

# This is a good example of a function.
# It takes in a parameter, and returns the result.
# The scope of the function all stays inside of the function, with no
# contact with the outside items.
add_one <- function(x) {
  x1 <- x + 1
  return(x1)
}
y <- 0
print(add_one(y))


# Note scope! Variables inside of our code don't last after the
# function finishes running
add_one <- function(x) {
  temporary_variable <- x + 1
  return(temporary_variable)
}
# The temporary_variable doesn't exist anymore!
add_one(1)


```

## Rules! 

There are some rules for creating and modifying functions.

For CS people, note that r uses pass-by-value. So, if you pass a variable, it won't be updated by the function. However, it's good practice to avoid updating a parameter value.

```{r eval= FALSE, echo=TRUE}

a <- 0

test_function <- function(a) {
  a <- 1
}
test_function(a)
print(a)

```

Note that this also applies to tibbles. This can cause issues if you have a very large dataset and use functions.

```{r eval= FALSE, echo=TRUE}
library(tidyverse)

a <- tibble(names = c('bob'))

test_function <- function(a) {
  a <- mutate(a, names = c('not bob'))
  print(a)
}
test_function(a)
print(a)

```


```{r eval=FALSE, echo=TRUE}


# Don't update the parameters. Instead, return the result.
add_one_bad <- function(x) {
  x <- x + 1
  return(x)
}
add_one_good <- function(x) {
  x_new <- x + 1
  return(x_new)
}


# Don't have side effects.
#
# It is possible to break the connection between inside of the function
# and outside.
#
# To do so, use <<- instead of <-.  The latter just creates a new local
# variable inside of the block of code.
#
# In general, this is a bad idea.
y <- 0
add_to_y <- function() {
  # This doesn't create a new variable, instead it updates y outside of the
  # function block.
  y <<- y + 1
}
add_to_y()


# Don't access variables outside of your function.
#
# This is legal, but a bad idea because it makes it harder
# to update and understand your code.
z <- 1
y <- 0

add_bad <- function(x) {
  return(x + z)
}
add_bad(y)

add_good <- function(x1, x2) {
  return(x1 + x2)
}
add_good(y, z)
```

