---
title: Special topics in R
output: 
  html_document:
    toc: true
    theme: united
---

# Date

Date/times in R require some special consideration.
They are a distinct type, and require the as.Date function to turn from text (i.e., '01-01-2022' into a date).

``` {r}

date_as_text <-  "2001-04-03"
date_as_date <-  as.Date(date_as_text)

print(class(date_as_date))


datetime_as_text <- "1970-01-01 11:22:00 UTC"
datetime_as_posix <- as.POSIXct(datetime_as_text)

print(class(datetime_as_posix))
```

## Storage

Dates are stored internally as the number of days before/after January 1, 1970. This can be negative or positive.

We store time as a fraction of a day. So noon would be 0.5.

``` {r}
dt <- as.Date('1950-01-01')
unclass(dt)
```

## Format function

We can output dates in different styles with the `format` command: 

- `%d` number day as `19`
- `%a` text weekday as `Sun` (or `%A` for `Sunday`)
- `%m` number month as `04`
- `%b` text month as `Feb` (or `%B` as `February`)
- `%y` year as `14` (`%Y` as `2014`)

``` {r}

d <- as.Date("2001-07-01")

vector_of_date_as_text_values <- c(
  format(d, format = "%d %m %y"),
  format(d, format = "%d-%m-%y"),
  format(d, format = "%B, %Y"),
  format(d, format = "%a (%d of %b)")
)
print(vector_of_date_as_text_values)
```

# Lubridate

We typically use lubridate in this class to simplify date/time handling. The biggest issue to keep straight is if an item is a date or a date-time. See below for examples.


``` {r}
library(lubridate)


d <- ymd("2023-02-03")
dt <- ymd_hm("2023-02-03 12:23")
```


# Cool articles on time

How do we deal with leap dates? What about leap seconds?
 
- [Leap seconds](https://arstechnica.com/science/2023/11/leap-seconds-could-become-leap-minutes-despite-pushback-from-russians-vatican/)
- [Where does time come from?](https://dotat.at/@/2023-05-26-whence-time.html)
